---
title: "Linking Climate and Health Data"
author: "HEYA Research Group"
site: bookdown::bookdown_site
documentclass: book
#bibliography: ["book.bib"]
link-citations: yes
description: "A bookdown template linking to our GitHub project. EDIT THIS LATER"
---

# Preface {#preface}

This work is supported by current and past members of the [HEYA Research Group](www.heyaresearch.ca):

Rebekah McKinnon PhD,  Ishwar Tiwari PhD,  Jayden Chang MSc,  Jack McIlraith MSc,  Gina Martin PhD

Welcome to this bookdown project. You can track issues and tasks here:

-   GitHub Project Board: [View on GitHub](https://github.com/RebekahMcKinnon/Climate_and_Health/projects/1)

## Introduction {#introduction}

### Purpose {#purpose}

## Required Information {#required-information}

### Datasets {#datasets}

This tutorial uses the following **publicly available** datasets:

| Data Provider                          | Dataset                                                                                                                                                                                                                              | Type                |
|-----------------------|------------------------------|-------------------|
| Natural Resources Canada (CFS / CWFIS) | National Burned Area Composite (**NBAC**) wildfire perimeters (1972--2024) ([cwfis.cfs.nrcan.gc.ca](<https://cwfis.cfs.nrcan.gc.ca/datamart/metadata/nbac?utm_source=chatgpt.com> "NBAC - Canadian Wildland Fire Information System  | Metadata"))         |
| Natural Resources Canada (EGS)         | **Floods in Canada -- Archive** (flood extent products) ([open.canada.ca](https://open.canada.ca/data/en/dataset/74144824-206e-4cea-9fb9-72925a128189?utm_source=chatgpt.com "Floods in Canada - Archive - Open Government Portal")) | Polygon (vector)    |
| Environment and Climate Change Canada  | Historical climate observations & station metadata ([climate.weather.gc.ca](%22https://climate.weather.gc.ca/climate_data/bulk_data_e.html?format=csv&stationID=%s&Year=%s&Month=1&Day=1&timeframe=2&submit=Download+Data%22))       | Point + time series |

 

This tutorial also uses datasets with **access restrictions**. Access may require institutional access, and/or data-use agreements.

| Data Provider                                                   | Dataset                                         | Access                                       |
|-------------------|-----------------------------|------------------------|
| Statistics Canada                                               | Postal Code Conversion File (**PCCF**) v8B      | Licensed                                     |
| Canadian Urban Environmental Health Research Consortium (CANUE) | Local Climate Zones & PM2.5 (smoke) by Postcode | Request via project creation in CANUE (free) |

### R packages {#r-packages}

The following R packages are required to run the code in this book:

**Data processing and wrangling**

-   `dplyr`
-   `tidyverse`\
-   `tidyr`
-   `lubridate`
-   `data.table`
-   `tibble`
-   `readr`
-   `purrr`
-   `stringr`
-   `readxl`
-   `writexl`

**Data visualization and mapping**

-   `ggplot2`
-   `maps`

**Spatial analysis**

-   `sf`
-   `geosphere`

**Data download**

-   `weathercan`

***Optional***

-   `furrr` (offers improved processing speeds)

The following code installs (if needed) and loads these packages:

```{r}
# Vector of required packages
packages <- c(
  "dplyr", "tidyr", "lubridate", "data.table", "tibble",
  "readr", "purrr", "stringr", "readxl", "writexl",
  "ggplot2", "maps",
  "sf", "geosphere",
  "weathercan", "furrr", "tidyverse"
)

# Install any that are missing
installed <- rownames(installed.packages())
to_install <- setdiff(packages, installed)
if (length(to_install) > 0) {
  install.packages(to_install)
}

# Load all packages
invisible(lapply(packages, library, character.only = TRUE))
```

<!--chapter:end:index.Rmd-->

# Getting Started with Spatial Data {#getting-started}

## Data Access {#data-access}

## Data preparation {#data-preparation}

### Terminology {#terminology}

### Coordinate Reference Systems & Projections {#coordinate-reference-systems-and-projections}



<!--chapter:end:chapters/01-getting-started.Rmd-->

# Spatial Data {#spatial-data}


<!--chapter:end:chapters/02-spatial-data.Rmd-->

# Locating & Refining Environmental Data {#locating-and-refining-environmental-data}

## Introduction {#introduction}

## Wildfires {#wildfires}

### Data preparation with sf package {#data-preparation-with-sf-package}

We read the updated NBAC wildfire polygons (1972–2024) from the .shp file.

- Tip: Shapefiles are a set of files. Ensure the .shp, .shx, .dbf, and (ideally) .prj sit together in the same folder.

```{r}
poly_fires <- st_read("C:/Users/rebek/Downloads/NBAC_1972to2024_20250506_shp/NBAC_1972to2024_20250506.shp") 

names(poly_fires)
```

### Inspect & refine data structure {#inspect-and-refine-data-structure}

Check available columns and data types 

```{r}
names(poly_fires)
str(poly_fires)
```

Normalise column names 

```{r}
names(poly_fires) <- tolower(names(poly_fires)) 
names(poly_fires)
```

Given the extremely large size, it's best to reduce the dataset to include only the years you require 

```{r}
poly_fires <- poly_fires %>% 
  filter(year >= 2010)
```

Limit dataset to only the columns you require 

```{r}
poly_fires <- select(poly_fires, 
                     nfireid, basrc,
                     year, 
                     ag_sdate, # start date as date format
                     ag_edate, # end date as date format
                     poly_ha, 
                     admin_area,
                     geometry)

head(poly_fires)
```

Rename columns to align better with your spatial datasets 

```{r}
poly_fires <- poly_fires %>%
  mutate(
    start_day = day(ag_sdate),
    start_month = month(ag_sdate),
    start_year = year(ag_sdate),
    end_day = day(ag_edate),
    end_month = month(ag_edate),
    end_year = year(ag_edate),
    start_date = ag_sdate, 
    end_date = ag_edate,
    size = poly_ha) %>%
  select(-poly_ha)  # remove the original poly_ha column


names(poly_fires)
str(poly_fires)
```

You may also want to save this newly formed dataset for easier future loading (i.e., to avoid the need to reload the shapefile during your next R session)

```{r}
# uncomment to run
# write.csv(poly_fires, "wildfire_data.csv", row.names = FALSE)
```


<!--chapter:end:chapters/03-locating-and-refining-environmental-data.Rmd-->

