---
title: "Linking Climate and Health Data"
author: "HEYA Research Group"
site: bookdown::bookdown_site
documentclass: book
#bibliography: ["book.bib"]
link-citations: yes
description: "A bookdown template linking to our GitHub project. EDIT THIS LATER"
---

# Preface {#preface}

This work is supported by current and past members of the [HEYA Research Group](www.heyaresearch.ca):

Rebekah McKinnon PhD,  Ishwar Tiwari PhD,  Jayden Chang MSc,  Jack McIlraith MSc,  Gina Martin PhD

Welcome to this bookdown project. You can track issues and tasks here:

-   GitHub Project Board: [View on GitHub](https://github.com/RebekahMcKinnon/Climate_and_Health/projects/1)

## Introduction {#introduction}

### Purpose {#purpose}

## Required Information {#required-information}

### Datasets {#datasets}

This tutorial uses the following **publicly available** datasets:

| Data Provider                          | Dataset                                                                                                                                                                                                                              | Type                |
|-----------------------|------------------------------|-------------------|
| Natural Resources Canada (CFS / CWFIS) | National Burned Area Composite (**NBAC**) wildfire perimeters (1972--2024) ([cwfis.cfs.nrcan.gc.ca](<https://cwfis.cfs.nrcan.gc.ca/datamart/metadata/nbac?utm_source=chatgpt.com> "NBAC - Canadian Wildland Fire Information System  | Metadata"))         |
| Natural Resources Canada (EGS)         | **Floods in Canada -- Archive** (flood extent products) ([open.canada.ca](https://open.canada.ca/data/en/dataset/74144824-206e-4cea-9fb9-72925a128189?utm_source=chatgpt.com "Floods in Canada - Archive - Open Government Portal")) | Polygon (vector)    |
| Environment and Climate Change Canada  | Historical climate observations & station metadata ([climate.weather.gc.ca](%22https://climate.weather.gc.ca/climate_data/bulk_data_e.html?format=csv&stationID=%s&Year=%s&Month=1&Day=1&timeframe=2&submit=Download+Data%22))       | Point + time series |

 

This tutorial also uses datasets with **access restrictions**. Access may require institutional access, and/or data-use agreements.

| Data Provider                                                   | Dataset                                         | Access                                       |
|-------------------|-----------------------------|------------------------|
| Statistics Canada                                               | Postal Code Conversion File (**PCCF**) v8B      | Licensed                                     |
| Canadian Urban Environmental Health Research Consortium (CANUE) | Local Climate Zones & PM2.5 (smoke) by Postcode | Request via project creation in CANUE (free) |

### R packages {#r-packages}

The following R packages are required to run the code in this book:

**Data processing and wrangling**

-   `dplyr`
-   `tidyverse`
-   `tidyr`
-   `lubridate`
-   `data.table`
-   `tibble`
-   `readr`
-   `purrr`
-   `stringr`
-   `readxl`
-   `writexl`
-   `rvest`
-   `httr`
-   `fs`
-   `lubridate`

**Data visualization and mapping**

-   `ggplot2`
-   `maps`

**Spatial analysis**

-   `sf`
-   `geosphere`

**Data download**

-   `weathercan`

***Optional***

-   `furrr` (offers improved processing speeds)

The following code installs (if needed) and loads these packages:

```{r}
# Vector of required packages
packages <- c(
  "dplyr", "tidyr", "lubridate", "data.table", "tibble",
  "readr", "purrr", "stringr", "readxl", "writexl",
  "ggplot2", "maps", "rvest", "httr", "fs", "lubridate",
  "sf", "geosphere",
  "weathercan", "furrr", "tidyverse"
)

# Install any that are missing
installed <- rownames(installed.packages())
to_install <- setdiff(packages, installed)
if (length(to_install) > 0) {
  install.packages(to_install, repos = c("https://ropensci.r-universe.dev", "https://cloud.r-project.org"))
}

# Load all packages
invisible(lapply(packages, library, character.only = TRUE))
```

<!--chapter:end:index.Rmd-->

# Getting Started with Spatial Data {#getting-started}

## Data Access {#data-access}

## Data preparation {#data-preparation}

### Terminology {#terminology}

### Coordinate Reference Systems & Projections {#coordinate-reference-systems-and-projections}

<!--chapter:end:chapters/01-getting-started.Rmd-->

# Spatial Data {#spatial-data}

<!--chapter:end:chapters/02-spatial-data.Rmd-->

# Locating & Refining Environmental Data {#locating-and-refining-environmental-data}

## Introduction {#introduction}

## Wildfires {#wildfires}

### Data preparation with sf package {#data-preparation-with-sf-package}

We read the updated NBAC wildfire polygons (1972-2024) from the .shp file.

- Tip: Shapefiles are a set of files. Ensure the .shp, .shx, .dbf, and (ideally) .prj sit together in the same folder.

```{r}
poly_fires <- st_read("C:/Users/rebek/Downloads/NBAC_1972to2024_20250506_shp/NBAC_1972to2024_20250506.shp") 

names(poly_fires)
```

### Inspect & refine data structure {#inspect-and-refine-data-structure}

Check available columns and data types

```{r}
names(poly_fires)
str(poly_fires)
```

Normalise column names

```{r}
names(poly_fires) <- tolower(names(poly_fires)) 
names(poly_fires)
```

Given the extremely large size, it's best to reduce the dataset to include only the years you require

```{r}
poly_fires <- poly_fires %>% 
  filter(year >= 2010)
```

Limit dataset to only the columns you require

```{r}
poly_fires <- select(poly_fires, 
                     nfireid, basrc,
                     year, 
                     ag_sdate, # start date as date format
                     ag_edate, # end date as date format
                     poly_ha, 
                     admin_area,
                     geometry)

head(poly_fires)
```

Rename columns to align better with your spatial datasets

```{r}
poly_fires <- poly_fires %>%
  mutate(
    start_day = day(ag_sdate),
    start_month = month(ag_sdate),
    start_year = year(ag_sdate),
    end_day = day(ag_edate),
    end_month = month(ag_edate),
    end_year = year(ag_edate),
    start_date = ag_sdate, 
    end_date = ag_edate,
    size = poly_ha) %>%
  select(-poly_ha)  # remove the original poly_ha column


names(poly_fires)
str(poly_fires)
```

You may also want to save this newly formed dataset for easier future loading (i.e., to avoid the need to reload the shapefile during your next R session)

```{r}
# uncomment to run
# write.csv(poly_fires, "wildfire_data.csv", row.names = FALSE)
```

## Hurricanes {#hurricanes}

### Data Sourcing

Hurricane data was sourced from the National Oceanic and Atmospheric Administration (NOAA). Data can be obtained programmatically (as below), or manually for each hurricane of interest from the [NOAA website](https://www.nhc.noaa.gov/gis/archive_besttrack.php?year= "Hurricane Best Tracks"). Should you choose to download the files yourself, the script is designed to use shapefiles (.shp).

- It is important to note that the following script should be used sparingly 


We first need to specify where we are getting the hurricane data from and for which years. 
```{r}
base_page <- "https://www.nhc.noaa.gov/gis/archive_besttrack.php?year="
base_zip  <- "https://www.nhc.noaa.gov/gis/best_track/"
hurricane_years_interest <- 2008:2025 #getting data for all years from 2008 to 2025
```

Next, we set up the storage location for the hurricane best track files and create a lookup table for later. It is important to set a viable output directory for the best tracks from the website. 
```{r}
#set an output directory for the data to download into
base_shape_dir <- "./HurricaneRawData/nhc_best_tracks"
dir.create(base_shape_dir, showWarnings = FALSE)

#create a lookup table for the storms that can be used later to create a connection between the storm name and data
storm_summary <- data.frame(Year = integer(), Storms = integer())
storm_lookup  <- data.frame(year = integer(), storm_id = character(), storm_name = character())
```

Finally, get the files from the website for each year specified. 

```{r, eval = FALSE}
#process every year for the year range specified above
for (year in hurricane_years_interest) {
  cat("processing", year, "\n")
  
  #read the page, catch if the response is blank
  page <- tryCatch({
    read_html(paste0(base_page, year))
  }, error = function(e) {
    cat("could not read page for", year, "\n")
    return(NULL)
  })
  if (is.null(page)) next
  
  #extract all the rows on the page to get the hurricane codes and names from the table for that year
  rows <- html_nodes(page, "tr")
  
  #loop through rows to match "al##" and extract name from second column to create lookup table
  for (row in rows) {
    cols <- html_nodes(row, "td") #get all the td html elements
    
    if (length(cols) >= 2) {
      id_text <- str_trim(html_text(cols[1])) #just want the id from the td elements
      
      if (str_detect(tolower(id_text), "^al\\d{2}$")) { #regex to get the atlantic (al) storm and the id number
        name_text <- str_trim(html_text(cols[2])) #get the nominal identifier
        storm_lookup <- rbind( #collect and add to the storm loookup table
          storm_lookup,
          data.frame( #create a data frame with the year, storm id from the td element, and the storm name
            year = year,
            storm_id = tolower(id_text),
            storm_name = name_text,
            stringsAsFactors = FALSE
          )
        )
      }
    }
  }
  
  #get just the text from all the td html elements
  td_text <- html_text(html_nodes(page, "td"), trim = TRUE)
  
  #get the strom ids again (for redundancies sake)
  storm_ids <- unique(str_to_lower(str_subset(td_text, "^al\\d{2}")))
  
  #make sure it's just the 4 character id with more regex
  storm_ids <- gsub("\\s.*", "", storm_ids)
  
  cat(" found", length(storm_ids), "atlantic storms\n")
  
  #storm summary for creating storm ids in file names
  storm_summary <- rbind(storm_summary,
                         data.frame(Year = year, Storms = length(storm_ids)))
  
  #file naming system is regular within the website hierarchy, so download all zip shapefiles using the engineered storm names
  for (storm_id in storm_ids) {
    zip_name <- paste0(storm_id, year, "_best_track.zip")
    url <- paste0(base_zip, zip_name)
    dest <- file.path(base_shape_dir, zip_name) #set outpath based on string specified earlier
    
    cat(" downloading:", zip_name, "\n")
    resp <- tryCatch({
      GET(url, write_disk(dest, overwrite = TRUE)) #get zip file from constructed url
    }, error = function(e) {
      cat(" failed to download", zip_name, "\n") #sometimes there are no files to download, especially in earlier years
      return(NULL)
    })
    
    #unzip the files too, easier to do now than later
    if (!is.null(resp) && status_code(resp) == 200) {
      unzip(dest, exdir = file.path(base_shape_dir, paste0(storm_id, "_", year)))
    } else {
      cat(" file not found:", url, "\n")
    }
    #be kind to the server and don't get banned from scraping
    Sys.sleep(1.1)
  }
}

write.csv(storm_lookup, r"(./HurricaneRawData/storm_lookup_table.csv)")
```


### Data Cleaning

Now that the data is downloaded and ready, it is time to clean the data up.

```{r}
#standardizing naming conventions in files
month_map <- c(
  "jan" = 1, "january" = 1, "01" = 1,
  "feb" = 2, "february" = 2, "02" = 2,
  "mar" = 3, "march" = 3, "03" = 3,
  "apr" = 4, "april" = 4, "04" = 4,
  "may" = 5, "05" = 5,
  "jun" = 6, "june" = 6, "06" = 6,
  "jul" = 7, "july" = 7, "07" = 7,
  "aug" = 8, "august" = 8, "08" = 8,
  "sep" = 9, "september" = 9, "09" = 9,
  "oct" = 10, "october" = 10, "10" = 10,
  "nov" = 11, "november" = 11, "11" = 11,
  "dec" = 12, "december" = 12, "12" = 12
)

#base directory where files are saved
base_dir <- r"(./HurricaneRawData/nhc_best_tracks)"

#set an ID for all of the files in the pts files and radii files
#get all shapefiles recursively
shp_files <- list.files(base_dir, pattern = "(_pts\\.shp$|_radii\\.shp$)", recursive = TRUE, full.names = TRUE)
```



### Data Manipulation


<!--chapter:end:chapters/03-locating-and-refining-environmental-data.Rmd-->
