```{r setup-introduction, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/matth/Climate_and_Health/bookdown_project')
```


# Case Study

In the following examples, we will be using a random selection of hospitals from the [Open Database of Hospital Facilities in Canada](#../index.Rmd/datasets "Datasets"). Hospitals frequently house at-risk populations of individuals, and are an interesting point to make a case study demonstrating this workflow. 

We are imagining that the hospital administration delivered a survey to all employees that had worked at each location for a minimum of two years. This data is then being analyzed by us, to determine the impact of environmental exposures in the past two years. 

```{r}
odhf_valid <- read.csv("./data/odhf_hospitals.csv", row.names = 1)#read in data

odhf_display <- odhf_valid %>% 
  select("lat","long","facility_name","postalcode","survey_month","survey_year") %>% #select desired columns
  st_as_sf(coords=c("long", "lat"), crs=st_crs(4326)) %>% #change to sf object
  mutate(across(where(is.character), ~ iconv(.x, from = "", to = "UTF-8", sub = ""))) #fix encoding problem

odhf_display <- st_transform(odhf_display, 3978)#projected coordinate system

mapview(odhf_display, layer.name="Hospitals")#show on map
```

