```{r setup-fires, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/matth/Climate_and_Health/bookdown_project')
```

# Wildfires

Wildfires can have strong impacts on an environment that are disproportionate to the footprint size of the fires themselves. Smoke from fires has been recorded to stretch thousands of kilometres #REF#. Wildfires are widely studied in environmental exposure impact analysis...

## Environmental Data Preparation {#fire-data-preparation}

Data can be downloaded for free from the [Canadian Wildland Fire Information System](#../index.Rmd/datasets "Datasets"). To begin with the wildfire data, we read the NBAC wildfire polygons (1972-2024) from the .shp file available from CWFIS.

-   Tip: Shapefiles are a set of files. Ensure the .shp, .shx, .dbf, and (ideally) .prj sit together in the same folder.

```{r fire-load-and-limit-data, eval=FALSE}
#read in the NBAC data
poly_fires <- st_read(r"(./data/FireData/NBAC_1972to2024_20250506.shp)") 

#look at the data
str(poly_fires)

#normalise column names
names(poly_fires) <- tolower(names(poly_fires)) 

#Given the extremely large size, it's best to reduce the dataset to include only the years required
poly_fires <- poly_fires %>% 
  filter(year >= 2015)

#limit to a selection of columns that are relevant
poly_fires <- select(poly_fires, 
                     nfireid, basrc,
                     year, 
                     ag_sdate, #agency start date as date format
                     ag_edate, #agency end date as date format
                     poly_ha, 
                     admin_area,
                     geometry)

#rename columns for easier understanding
poly_fires <- poly_fires %>%
  mutate(
    start_day = day(ag_sdate),
    start_month = month(ag_sdate),
    start_year = year(ag_sdate),
    end_day = day(ag_edate),
    end_month = month(ag_edate),
    end_year = year(ag_edate),
    start_date = ag_sdate, 
    end_date = ag_edate,
    size = poly_ha) %>%
  select(-poly_ha)  #remove the original poly_ha column

#write the newly formed dataset to memory, it will be loaded and used in this example
st_write_parquet(poly_fires, r"(./data/wildfire_data_sf.parquet)")

names(poly_fires)
```
```{r fire-data-pickup, include = FALSE}
#read the fire data snippet for reduced memory constraints
poly_fires <- st_read_parquet(r"(./data/wildfire_data_sf.parquet)")
```

## Compute Intersections {#fire-intersections}

Here, we load the hospital data for computing intersections. We are going to add a 25km buffer around each hospital to help identify fires with a perimeter that approached the hospitals.

```{r load-hospital-data}
buffer_distance = 25000 #in metres

#load hospital data
odhf_data <- read.csv("./data/odhf_hospitals.csv", row.names = 1)

# get ready to convert hospital data
odhf_fires <- odhf_data %>% 
  mutate(long = as.numeric(long)) %>% 
  filter(!is.na(lat) & !is.na(long)) #filter out na points

odhf_fires_sf <- odhf_fires %>% 
  st_as_sf(coords=c("long", "lat"), crs = st_crs(4326))

#transform to a metres pcs
odhf_fires_sf <- st_transform(odhf_fires_sf, st_crs(poly_fires)) 

#Create buffer around each location
odhf_buffers <- st_buffer(odhf_fires_sf, dist = buffer_distance) #buffer in metres
```

We need to compute the intersections between the fires and buffered hospitals. To streamline this example, we select only fires greater than 200 hectares for computing intersections using EPSG 3978. 

```{r fire-odhf-intersection}
min_fire_size <- 200 #in ha

#filter fires by size
poly_fires <- poly_fires %>% filter(size > min_fire_size)

#intersection matrix, the product refers to row indices, not id numbers
fire_matches <- st_intersects(odhf_buffers, poly_fires) #run st_intersection to get geometries, but takes way longer
```

Now that we have found the indexes of the intersections in `fire_matches`, we need to link this information back to the original, more detailed dataframe. 

```{r fire-link-intersections}
#create the intersection dataframe from the intersection matrix
intersection_df <- map2_dfr(seq_along(fire_matches),  fire_matches,
  ~ tibble(
    hospital_id = odhf_fires_sf$hospital_id[.x], #.x indicates it's from the sequence (argument 1)
    n_fire_id = poly_fires$nfireid[.y], #.y indicates it's from the fires (argument 2)
    survey_year = odhf_fires_sf$survey_year[.x],
    survey_month = odhf_fires_sf$survey_month[.x],
    fire_start_date = poly_fires$ag_sdate[.y],
    fire_end_date = poly_fires$ag_edate[.y],
    geometry = poly_fires$geometry[.y],
    intersects = 1
  )
)

#create useful date fields for the 24-month date lag
intersection_df$survey_eom_date <- ymd(paste(intersection_df$survey_year, intersection_df$survey_month, "01", sep = "-")) + months(1) - days(1)
intersection_df$survey_25_prev <- ymd(paste(intersection_df$survey_year, intersection_df$survey_month, "01", sep = "-")) -months(24)
intersection_df$fire_start_date <- as.Date(intersection_df$fire_start_date, format="%YYYY-%MM-%DD")#make into a date
intersection_df$fire_end_date <- as.Date(intersection_df$fire_end_date, format="%YYYY-%MM-%DD")#make into a date
```

We also need to deal with extraneous cases like rows with no fire end dates and those that were recorded as instantaneous (same start and end date).

```{r fire-date-range-coverage}
#this gets all overlapping dates, treating rows with no fire end date as one day instantaneous events
intersection_overlap <- intersection_df %>%
  mutate(
    #define effective fire range: point if no end date, range otherwise
    fire_start_eff = fire_start_date,
    fire_end_eff   = coalesce(fire_end_date, fire_start_date)
  ) %>%
  filter(
    #survey range overlaps with fire range
    survey_25_prev <= fire_end_eff,
    survey_eom_date >= fire_start_eff
  )
```

## Date Lag {#fire-date-lag}

As part of the exposure determination, we are interested in the 24 months prior to each survey date at the hospitals. Hospitals were surveyed at different times, so the date of the survey will be recorded along with each month from the 0th to 24th month prior (survey month, survey month - 1, survey month - 2, etc.).

```{r fire-date-lag, message=FALSE}
#expand each survey row into last 24 months
expand_months <- function(d, hospital_id, survey_month, survey_year) {
  data.table(
    hospital_id = hospital_id,
    survey_year = survey_year,
    survey_month = survey_month,
    offset = 0:24,
    date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
  )
}

#apply the function
#identifies how many months a given month is offset from the 0th month for that hospitals survey date
odhf_fires_sf$dates <- ymd(paste(odhf_fires_sf$survey_year, odhf_fires_sf$survey_month, "01", sep = "-")) + months(1) - days(1)
#for unique examples (one hospital-survey combo)
fires_un <- unique(odhf_fires_sf, by = c("hospial_id","dates"))
fires_un <- as.data.table(fires_un)
survey_expanded <- fires_un[, expand_months(dates, hospital_id, survey_month, survey_year), by = .(hospital_id, dates)] #expand the months backwards for each unique hospital_id/date
survey_expanded <- survey_expanded %>% select(-3) #get rid of the duplicate column
```

## Binary Matrix {#fire-summary-matrix}

Create a binary matrix for each hospital-survey date combination for each of the 24 months.

```{r fire-summary-matrix, warning=FALSE}
#create a detailed data frame of all intersections and non-intersections, regardless of date, as an intermediary product 
fires_detailed <- survey_expanded %>%
  left_join(intersection_df, by = c("hospital_id", "survey_year", "survey_month")) %>% #join the data on year month and id
  mutate(
    month_ym = ymd(paste0(date, "01")), #set the date
    month_start = month_ym, #get the first month as a date
    month_end = ceiling_date(month_ym, "month") - days(1), #get the last month as a date
    fire_start_eff = fire_start_date, #all have start date or nothing
    fire_end_eff = coalesce(fire_end_date, fire_start_date), #coalesce if nothing 
    intersect = case_when(
      is.na(fire_start_date) ~ 0,  #fire had no date or there is no fire
      month_start <= fire_end_eff & month_end >= fire_start_eff ~ 1, 
      TRUE ~ 0
    )
  ) %>%
  select(hospital_id, survey_year, survey_month, offset, date, #select these columns
         intersect, n_fire_id, fire_start_date, fire_end_date, fire_start_eff, fire_end_eff, geometry)

#create binary matrix of occurrences
fires_summary_matrix <- fires_detailed %>% 
  #filter(intersect == 1) %>% #if only interested in the results that have intersections
  group_by(hospital_id, survey_year, survey_month, offset) %>%  #group by hospital id
  right_join(survey_expanded, by=c("hospital_id", "survey_year", "survey_month", "offset")) %>% #join
  mutate(intersect = replace_na(intersect, 0)) %>% #make sure there are no NAs
  ungroup() %>%
  as.data.table() %>% #convert to data table for some easier mutability
  dcast(., #reshape data to wide format based on preceding months
      hospital_id + survey_month + survey_year ~ paste0("month_", offset),
      value.var = "intersect",
      fun.aggregate = max
    ) %>%
  setcolorder(c("hospital_id", "survey_month", "survey_year", paste0("month_", 0:24))) #column order makes sense 
  
  
#create an output sample to display
fires_summary_matrix[c(120:130), ] %>% 
  kbl(format = "html", caption = "Fire Summary Matrix") %>%
  kable_styling() %>%
  scroll_box(width = "100%")

#write csv result
write.csv(fires_summary_matrix, r"(./data/fires_presence_absence.csv)")

```

## Map {#fire-map}

The South Okanagan General Hospital in British Columbia experiences every exposure illustrated in this bookdown project except for hurricane events. Below is a visual depiction of how wildfire events interact with the hospital and the areas near it. 

```{r fire-map}
fires_summary_matrix <- read.csv(r"(./data/fires_presence_absence.csv)", row.names=1)

#join matrix back with hospital ids
fires_summary_map <- fires_summary_matrix %>% 
  left_join(odhf_fires_sf %>% select(hospital_id, geometry), by = "hospital_id")

#get intersecting fires
hospital_509 <- fires_detailed %>% 
  filter(hospital_id == 509, intersect == 1)

#display map elements
mapview(st_as_sf(hospital_509), layer.name="Intersecting Fires") + mapview(filter(odhf_fires_sf, hospital_id == 509), col.regions='red', cex=6, layer.name = "Hospital Location") + 
mapview(
  st_buffer(filter(odhf_fires_sf, hospital_id == 509), buffer_distance),
  col.regions = "green",
  alpha.regions = 0.1,
  layer.name = "Hospital Buffer (25km)"
)

```








