```{r setup-floods, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/matth/Climate_and_Health/bookdown_project')
```

# Floods {#floods}

Floods are a common environmental hazard in Canada. Numerous flooding events have been reported in popular media and studied in literature...

## Environmental Data Preparation {#floods-data-preparation}

Natural Resources Canada provides a flood archive data set through [Open Canada](#../index.Rmd/datasets "Datasets"). The data is provided in bulk as a .gdb file with several layers. We are specifically interested in the 'EGS_Flood_Product_Archive' layer.

```{r floods-initial-data, warning=FALSE, results='hide'}
#load raw flood data
gdb_path <- "./data/EGS_Flood_Product_Archive.gdb"
st_layers(gdb_path)
flood_data <- st_read(gdb_path, layer = "EGS_Flood_Product_Archive", options = "METHOD=SKIP")
```

## Data Preparation {#floods-data-cleaning}

Before any analysis can proceed with the flood data, the hospital data needs to be loaded and converted to an appropriate coordinate reference system that aligns with the flood data. 

```{r floods-survey-data-preprocess}
#load the hospital data
odhf_data <- read.csv("./data/odhf_hospitals.csv", row.names = 1)
#need to make some data to verify
#names throw an error due to character incompatibility, so remove altogether
odhf_valid <- odhf_data
survey_sf <- odhf_valid %>% st_as_sf(coords=c("long", "lat"), crs=st_crs(4326)) %>% select(-facility_name) %>% st_transform(st_crs(flood_data))
```

## Compute Intersections {#floods-data-manipulation}

The following code loops through each hospital to determine flood exposure for the 24 months leading up to the month the survey was conducted at each location. We will use a 25km buffer to encapsulate floods that occurred in areas near to the hospitals. 

```{r floods-main-loop, message=FALSE, results='hide'}
buffer_distance = 25000 #in metres

#create a buffer around each poi
survey_buffered <- st_buffer(survey_sf, dist = buffer_distance)

#find floods within each poi's buffer
flood_matches <- st_intersects(survey_buffered, flood_data)

#extract year and month from flood event datetime
flood_data <- flood_data %>%
  mutate(flood_year = year(datetime_utc),
         flood_month = month(datetime_utc))

#initialize an empty data frame to store results
exposure_matrix <- matrix(0, nrow = nrow(survey_sf), ncol = 25)
colnames(exposure_matrix) <- paste0("month_", 0:24)  #column names for each month

#loop through each poi and determine flood exposure
for (i in seq_len(nrow(survey_sf))) {
  #extract the survey date from the survey related columns
  survey_year <- survey_sf$survey_year[i]
  survey_month <- survey_sf$survey_month[i]
  
  #compute the 24-month range
  month_seq <- seq.Date(from = as.Date(sprintf("%d-%02d-01", survey_year, survey_month)) %m-% months(1), 
                        length.out = 24, by = "-1 month")
  month_seq_df <- data.frame(year = year(month_seq), month = month(month_seq))
  
  #get the indices of intersecting floods for this poi
  flood_indices <- flood_matches[[i]]
  
  if (length(flood_indices) > 0) {  #if floods exist within the buffer
    flood_events <- flood_data[flood_indices, c("flood_year", "flood_month")]
    
    #check for each of the past 24 months if a flood occurred
    for (j in 1:24) {
      target_year <- month_seq_df$year[j]
      target_month <- month_seq_df$month[j]
      
      #if a flood occurred in this year-month, set exposure to 1
      if (any(flood_events$flood_year == target_year & flood_events$flood_month == target_month)) {
        exposure_matrix[i, j] <- 1
      }
    }
  }
}

#convert matrix to a data frame and merge with survey data
exposure_df <- as.data.frame(exposure_matrix)

#bind to the presence absence matrix
survey_floods_pa <- cbind(survey_sf, exposure_df)

#summarise based on the number of unique hospitals
survey_floods_pa %>% summarise(n_unique = n_distinct(hospital_id)) 

#create a summary dataset of hospitals with only flood events for easier viewing
survey_floods_pa_filtered <- survey_floods_pa %>%
  filter(if_any(starts_with("month_"), ~ . == 1))

#write to a directory 
write.csv(survey_floods_pa, r"(./data/flood_presence_absence.csv)")
```
Display a snippet of the floods matrix.

```{r floods-matrix-snippet}
#print a selection of data
kable(head(survey_floods_pa_filtered),'html') %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(height = "300px")
```

## Map {#floods-map}

The South Okanagan General Hospital in British Columbia borders the Okanagan River, which is prone to flooding. Below is a visual depiction of flooding events in the area adjacent to the hospital. 

``` {r floods-map, warning=FALSE}
floods_summary_matrix <- read.csv(r"(./data/flood_presence_absence.csv)", row.names=1)

survey_sf$hospital_id <- as.character(survey_sf$hospital_id)
floods_summary_matrix$hospital_id <- as.character(floods_summary_matrix$hospital_id)

#South Okanagan General Hospital
floods_summary_map <- floods_summary_matrix %>% 
  left_join(survey_sf %>% select(hospital_id, geometry), by = "hospital_id")

#get the associated floods
hospital_509 <- survey_floods_pa %>% 
  filter(hospital_id == 509)

#get the intersecting floods
flood_509 <- st_intersection(st_buffer(hospital_509, dist = buffer_distance), flood_data)

#generate the map elements
mapview(
  flood_509,
  layer.name = "Intersecting Floods",
  popup = leafpop::popupTable(
    flood_509,
    zcol = c("swath_id", "product_file_name")
  )
) +
mapview(
  filter(survey_sf, hospital_id == 509),
  col.regions = "red",
  cex = 6,
  layer.name = "Hospital Location"
) +
mapview(
  st_buffer(filter(survey_sf, hospital_id == 509), buffer_distance),
  col.regions = "green",
  alpha.regions = 0.1,
  layer.name = "Hospital Buffer (25km)"
)
```