# Locating & Refining Environmental Data {#locating-and-refining-environmental-data}

## Introduction {#introduction}

## Wildfires {#wildfires}

### Data preparation with sf package {#data-preparation-with-sf-package}

We read the updated NBAC wildfire polygons (1972–2024) from the .shp file.

- Tip: Shapefiles are a set of files. Ensure the .shp, .shx, .dbf, and (ideally) .prj sit together in the same folder.

```{r}
poly_fires <- st_read("C:/Users/rebek/Downloads/NBAC_1972to2024_20250506_shp/NBAC_1972to2024_20250506.shp") 

names(poly_fires)
```

### Inspect & refine data structure {#inspect-and-refine-data-structure}

Check available columns and data types 

```{r}
names(poly_fires)
str(poly_fires)
```

Normalise column names 

```{r}
names(poly_fires) <- tolower(names(poly_fires)) 
names(poly_fires)
```

Given the extremely large size, it's best to reduce the dataset to include only the years you require 

```{r}
poly_fires <- poly_fires %>% 
  filter(year >= 2010)
```

Limit dataset to only the columns you require 

```{r}
poly_fires <- select(poly_fires, 
                     nfireid, basrc,
                     year, 
                     ag_sdate, # start date as date format
                     ag_edate, # end date as date format
                     poly_ha, 
                     admin_area,
                     geometry)

head(poly_fires)
```

Rename columns to align better with your spatial datasets 

```{r}
poly_fires <- poly_fires %>%
  mutate(
    start_day = day(ag_sdate),
    start_month = month(ag_sdate),
    start_year = year(ag_sdate),
    end_day = day(ag_edate),
    end_month = month(ag_edate),
    end_year = year(ag_edate),
    start_date = ag_sdate, 
    end_date = ag_edate,
    size = poly_ha) %>%
  select(-poly_ha)  # remove the original poly_ha column


names(poly_fires)
str(poly_fires)
```

You may also want to save this newly formed dataset for easier future loading (i.e., to avoid the need to reload the shapefile during your next R session)

```{r}
write.csv(poly_fires, "wildfire_data.csv", row.names = FALSE)
```

