```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/matth/Climate_and_Health/bookdown_project')
```
# Locating & Refining Environmental Data {#locating-and-refining-environmental-data}

## Introduction {#introduction-environmental-data}

Sample data is Canadian public hospital data.

## Wildfires {#wildfires}

### Data preparation with sf package {#data-preparation-with-sf-package}

We read the updated NBAC wildfire polygons (1972-2024) from the .shp file.

-   Tip: Shapefiles are a set of files. Ensure the .shp, .shx, .dbf, and (ideally) .prj sit together in the same folder.

```{r load-fire-data}
#TODO create unified language and narrative view in html text throughout
poly_fires <- st_read(r"(./data/FireData/NBAC_1972to2024_20250506.shp)") 

names(poly_fires)
```

### Inspect & refine data structure {#inspect-and-refine-data-structure}

Check available columns and data types.

```{r check-fire-data}
names(poly_fires)
str(poly_fires)
```

Normalise column names.

```{r normalise-fire-data}
names(poly_fires) <- tolower(names(poly_fires)) 
names(poly_fires)
```

Given the extremely large size, it's best to reduce the dataset to include only the years required.

```{r filter-fire-data}
poly_fires <- poly_fires %>% 
  filter(year >= 2010)
```

Limit dataset to only the required columns.

```{r limit-fire-data}
poly_fires <- select(poly_fires, 
                     nfireid, basrc,
                     year, 
                     ag_sdate, # start date as date format
                     ag_edate, # end date as date format
                     poly_ha, 
                     admin_area,
                     geometry)

head(poly_fires)
```

Rename columns to align better with the spatial datasets.

```{r rename-fire-data}
poly_fires <- poly_fires %>%
  mutate(
    start_day = day(ag_sdate),
    start_month = month(ag_sdate),
    start_year = year(ag_sdate),
    end_day = day(ag_edate),
    end_month = month(ag_edate),
    end_year = year(ag_edate),
    start_date = ag_sdate, 
    end_date = ag_edate,
    size = poly_ha) %>%
  select(-poly_ha)  # remove the original poly_ha column


names(poly_fires)
str(poly_fires)
```

You may want this newly formed dataset saved for easier future loading (i.e., to avoid the need to reload the shapefile during the next R session).

```{r write-fire-data}
# uncomment to run
# write.csv(poly_fires, "wildfire_data.csv", row.names = FALSE)
```

## Floods {#floods}

### Data Sourcing {#floods-data-sourcing}

See the index.

```{r floods-initial-data, warning=FALSE}
# load data
gdb_path <- "./data/EGS_Flood_Product_Archive.gdb"
st_layers(gdb_path)
flood_data <- st_read(gdb_path, layer = "EGS_Flood_Product_Archive", options = "METHOD=SKIP")
kable(head(flood_data), format = "html") %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(height = "300px")
```

### Data Cleaning

We are interested in floods that happened within 25km of the POI survey data. After converting the survey data to the same coordinate reference system as the flood data we can look at the intersection of the two data points.

```{r floods-survey-data-preprocess}
# need to make some data to verify
# names throw an error due to character incompatibility, so remove altogether
survey_sf <- odhf_valid %>% st_as_sf(coords=c("long", "lat"), crs=st_crs(4326)) %>% select(-facility_name) %>% st_transform(st_crs(flood_data))
```

### Data Manipulation {#floods-data-manipulation}

Buffer distance is easily customisable for different use cases. This chunk will loop through each of the points of interest to determine flood exposure with the floods dataset. 

```{r floods-main-loop}
buffer_distance = 25000 # in metres

# create a buffer around each poi
survey_buffered <- st_buffer(survey_sf, dist = buffer_distance)

# find floods within each poi's buffer
flood_matches <- st_intersects(survey_buffered, flood_data)

# extract year and month from flood event datetime
flood_data <- flood_data %>%
  mutate(flood_year = year(datetime_utc),
         flood_month = month(datetime_utc))

# initialize an empty data frame to store results
exposure_matrix <- matrix(0, nrow = nrow(survey_sf), ncol = 25)
colnames(exposure_matrix) <- paste0("month_", 0:24)  #column names for each month

# loop through each poi and determine flood exposure
for (i in seq_len(nrow(survey_sf))) {
  # extract the survey date from the survey related columns
  survey_year <- survey_sf$survey_year[i]
  survey_month <- survey_sf$survey_month[i]
  
  # compute the 24-month range
  month_seq <- seq.Date(from = as.Date(sprintf("%d-%02d-01", survey_year, survey_month)) %m-% months(1), 
                        length.out = 24, by = "-1 month")
  month_seq_df <- data.frame(year = year(month_seq), month = month(month_seq))
  
  # get the indices of intersecting floods for this poi
  flood_indices <- flood_matches[[i]]
  
  if (length(flood_indices) > 0) {  # if floods exist in the buffer
    flood_events <- flood_data[flood_indices, c("flood_year", "flood_month")]
    
    # check for each of the past 24 months if a flood occurred
    for (j in 1:24) {
      target_year <- month_seq_df$year[j]
      target_month <- month_seq_df$month[j]
      
      # if a flood occurred in this year-month, set exposure to 1
      if (any(flood_events$flood_year == target_year & flood_events$flood_month == target_month)) {
        exposure_matrix[i, j] <- 1
      }
    }
  }
}

# convert matrix to a data frame and merge with survey data
exposure_df <- as.data.frame(exposure_matrix)
# bind in the presence absence matrix
survey_floods_pa <- cbind(survey_sf, exposure_df)

# summarise based on the number of unique hospitals
survey_floods_pa %>% summarise(n_unique = n_distinct(hospital_id)) 

# create a summary dataset of hospitals with only flood events for easier viewing
survey_floods_pa_filtered <- survey_floods_pa %>%
  filter(if_any(starts_with("month_"), ~ . == 1))

kable(head(survey_floods_pa_filtered),'html') %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(height = "300px")

# write to a directory 
#write.csv(survey_floods_pa, "flood_presence_absence_2yr.csv", row.names = FALSE)
```

## Hurricanes {#hurricanes}

### Data Sourcing {#hurricanes-data-sourcing}

Hurricane data was sourced from the National Oceanic and Atmospheric Administration (NOAA). Data can be obtained programmatically (as below), or manually for each hurricane of interest from the [NOAA website](https://www.nhc.noaa.gov/gis/archive_besttrack.php?year= "Hurricane Best Tracks"). Should you choose to download the files yourself, the script is designed to use shapefiles (.shp).

-   It is important to note that the following script should be used sparingly

We first need to specify when and where the hurricane data is retrieved from.

```{r hurricanes-set-data-extraction-range}
base_page <- "https://www.nhc.noaa.gov/gis/archive_besttrack.php?year="
base_zip  <- "https://www.nhc.noaa.gov/gis/best_track/"
hurricane_years_interest <- 2008:2025 #getting data for all years from 2008 to 2025
```

Next, we set up the storage location for the hurricane best track files and create a lookup table for later. It is important to set a viable output directory for the best tracks from the website.

```{r hurricanes-define-output}
#set an output directory for the data to download into
base_shape_dir <- "./data/HurricaneRawData/nhc_best_tracks"
dir.create(base_shape_dir, showWarnings = FALSE)

#create a lookup table for the storms that can be used later to create a connection between the storm name and data
storm_summary <- data.frame(Year = integer(), Storms = integer())
storm_lookup  <- data.frame(year = integer(), storm_id = character(), storm_name = character())
```

Finally, we get the files from the website for each year specified. Within this snippet, a lookup table for the file names and actual hurricane name is created and saved as a .csv for later reference. (TODO: add a switch to get just the names list vs. the file and names list)

```{r hurricanes-scrape-data, eval = FALSE}
#process every year for the year range specified above
for (year in hurricane_years_interest) {
  cat("processing", year, "\n")
  
  #read the page, catch if the response is blank
  page <- tryCatch({
    read_html(paste0(base_page, year))
  }, error = function(e) {
    cat("could not read page for", year, "\n")
    return(NULL)
  })
  if (is.null(page)) next
  
  #extract all the rows on the page to get the hurricane codes and names from the table for that year
  rows <- html_nodes(page, "tr")
  
  #loop through rows to match "al##" and extract name from second column to create lookup table
  for (row in rows) {
    cols <- html_nodes(row, "td") #get all the td html elements
    
    if (length(cols) >= 2) {
      id_text <- str_trim(html_text(cols[1])) #just want the id from the td elements
      
      if (str_detect(tolower(id_text), "^al\\d{2}$")) { #regex to get the atlantic (al) storm and the id number
        name_text <- str_trim(html_text(cols[2])) #get the nominal identifier
        storm_lookup <- rbind( #collect and add to the storm loookup table
          storm_lookup,
          data.frame( #create a data frame with the year, storm id from the td element, and the storm name
            year = year,
            storm_id = tolower(id_text),
            storm_name = name_text,
            stringsAsFactors = FALSE
          )
        )
      }
    }
  }
  
  #get just the text from all the td html elements
  td_text <- html_text(html_nodes(page, "td"), trim = TRUE)
  
  #get the strom ids again (for redundancies sake)
  storm_ids <- unique(str_to_lower(str_subset(td_text, "^al\\d{2}")))
  
  #make sure it's just the 4 character id with more regex
  storm_ids <- gsub("\\s.*", "", storm_ids)
  
  cat(" found", length(storm_ids), "atlantic storms\n")
  
  #storm summary for creating storm ids in file names
  storm_summary <- rbind(storm_summary,
                         data.frame(Year = year, Storms = length(storm_ids)))
  
  #file naming system is regular within the website hierarchy, so download all zip shapefiles using the engineered storm names
  for (storm_id in storm_ids) {
    zip_name <- paste0(storm_id, year, "_best_track.zip")
    url <- paste0(base_zip, zip_name)
    dest <- file.path(base_shape_dir, zip_name) #set outpath based on string specified earlier
    
    cat(" downloading:", zip_name, "\n")
    resp <- tryCatch({
      GET(url, write_disk(dest, overwrite = TRUE)) #get zip file from constructed url
    }, error = function(e) {
      cat(" failed to download", zip_name, "\n") #sometimes there are no files to download, especially in earlier years
      return(NULL)
    })
    
    #unzip the files too, easier to do now than later
    if (!is.null(resp) && status_code(resp) == 200) {
      unzip(dest, exdir = file.path(base_shape_dir, paste0(storm_id, "_", year)))
    } else {
      cat(" file not found:", url, "\n")
    }
    #be kind to the server and don't get banned from scraping
    Sys.sleep(1.1)
  }
}

write.csv(storm_lookup, r"(./data/HurricaneRawData/storm_lookup_table.csv)")
```

### Data Cleaning {#hurricanes-data-cleaning}

Now that the data is downloaded and ready, it is time clean it up. There are two files of interest for the hurricanes, the point files and the polygon (radii) files. These files need to be manipulated so that dates and ids are consistent between the two file sets and between years. 

```{r hurricanes-map-months}
#standardizing naming conventions in files
month_map <- c(
  "jan" = 1, "january" = 1, "01" = 1,
  "feb" = 2, "february" = 2, "02" = 2,
  "mar" = 3, "march" = 3, "03" = 3,
  "apr" = 4, "april" = 4, "04" = 4,
  "may" = 5, "05" = 5,
  "jun" = 6, "june" = 6, "06" = 6,
  "jul" = 7, "july" = 7, "07" = 7,
  "aug" = 8, "august" = 8, "08" = 8,
  "sep" = 9, "september" = 9, "09" = 9,
  "oct" = 10, "october" = 10, "10" = 10,
  "nov" = 11, "november" = 11, "11" = 11,
  "dec" = 12, "december" = 12, "12" = 12
)

#base directory where files are saved
base_dir <- r"(./data/HurricaneRawData/nhc_best_tracks)"

#set an ID for all of the files in the pts files and radii files
#get all shapefiles recursively
shp_files <- list.files(base_dir, pattern = "(_pts\\.shp$|_radii\\.shp$)", recursive = TRUE, full.names = TRUE)
```

Add the storm ID for each storm to the file names and overwrite the shapefile.

```{r hurricanes-add-storm-id, warning=FALSE, message=FALSE, results='hide'}

#process the files
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  #read the shapefile
  shp <- st_read(shp_path, quiet = TRUE)
  
  #extract part before first underscore
  file_prefix <- sub("_.*", "", basename(shp_path))
  
  #add the storm id file
  shp <- shp %>% mutate(storm_id = file_prefix)
  
  #overwrite shapefile
  st_write(shp, shp_path, delete_dsn = TRUE, quiet = TRUE)
}

```


Date and time need to be consistent between the point and polygon files.

```{r hurricanes-alter-points-files, message=FALSE, warning=FALSE, results='hide'}
#Create a new directory for processed files
output_dir <- file.path(base_dir, "processed")

#processing for just the _pts.shp files to normalize dates and times
base_dir <- r"(./data/HurricaneRawData/nhc_best_tracks)"
output_dir <- file.path(base_dir, "processed")
dir.create(output_dir, showWarnings = FALSE)

#get all the shapefiles recursively
shp_files <- list.files(base_dir, pattern = "_pts\\.shp$", recursive = TRUE, full.names = TRUE)

#recursively alter the pts files to the normalized date and time fields with ids
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  shp <- st_read(shp_path, quiet = TRUE)
  
  #files had different formats for dates and times, this normalizes them
  # important step to combine everything for later analysis
  shp <- shp %>%
    mutate(
      mm = str_pad(as.integer(unname(month_map[tolower(MONTH)])), 2, pad = "0"),
      dd = str_pad(as.integer(DAY), 2, pad = "0"),
      hh = str_sub(HHMM, 1, 2),
      event_id = tolower(paste0(storm_id, mm, dd, hh))
    )
  
  #preserve folder structure in output
  rel_path <- path_rel(shp_path, start = base_dir)
  rel_path <- sub("^processed/", "", rel_path)
  out_path <- file.path(output_dir, rel_path)
  dir_create(dirname(out_path))
  
  #write output
  st_write(shp, out_path, delete_dsn = TRUE, quiet = TRUE)
}
```


The points files also need to be done in a similar way.

```{r hurricanes-alter-radii-files, message=FALSE, warning=FALSE, results='hide'}
#slightly different process for the _radii file they already have the synoptime field and those just need to be modified
shp_files <- list.files(base_dir, pattern = "_radii\\.shp$", recursive = TRUE, full.names = TRUE)

#recursively alter the pts files to fie normalized date and time fields with ids
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  shp <- st_read(shp_path, quiet = TRUE)
  
  #create matching event_id to the points fields so they can be related (since they have different kinds of information)
  shp <- shp %>%
    mutate(
      event_id = tolower(
        paste0(storm_id, substr(SYNOPTIME, 5, nchar(SYNOPTIME))) #this will be storm id, date, and UTC hour in 24 hour format i.e. [al01][20250831][12]
      )
    )
  
  #different kind of id, radii refers to the range of wind at that knot level i.e. _34 would be 34 knot wind
  shp <- shp %>%
    mutate(
      event_r_id = tolower(
        paste0(storm_id, substr(SYNOPTIME, 5, nchar(SYNOPTIME)), "_", RADII)#this will be storm id, date, and UTC hour in 24 hour format, and radii distance  i.e. [al01][20250831][12][_34]
      )
    )
  
  #preserve folder structure
  rel_path <- path_rel(shp_path, start = base_dir)
  out_path <- file.path(output_dir, rel_path)
  dir_create(dirname(out_path))
  
  #write output
  st_write(shp, out_path, delete_dsn = TRUE, quiet = TRUE)
}
```


### Data Manipulation {#hurricanes-data-manipulation}

Now that the hurricane files are prepared properly, the associated points data of interest, (in this case hospitals), can be used within the files. The point files in this example have a series of months and years in columns when surveys were conducted, a latitude and longitude column where we can identify

First we load all the storms.

```{r hurricanes-load-storms}
#where all the processed files are
storms_dir <- r"(./data/HurricaneRawData/nhc_best_tracks/processed)"

#get all the radii shapefiles
#I settled on radii, because its the reach of the wind
shp_files <- list.files(storms_dir, pattern = "_radii\\.shp$", recursive = TRUE, full.names = TRUE)

#read and merge them together
storms <- lapply(shp_files, function(shp) {
  suppressWarnings(
    st_read(shp, quiet = TRUE) %>% 
      select(any_of(cols)) %>%
      st_transform(4326) %>%
      st_make_valid()
  )
})

the_storms <- do.call(rbind, storms)

all_storms <- the_storms

#make sure the geometries are valid, often times they were unclosed fan shapes
all_storms <- st_transform(all_storms, 4326)
all_storms <- st_make_valid(all_storms)
#get just the date from the storms
all_storms$months <- ym(str_sub(all_storms$event_id,5,10))
```

Now that all the data is loaded and ready, the intersection with all the points of interest needs to be performed and a number of data outputs are created.

1. A long table of storms and intersections with schools.
2. A wide table with the ID's of the surveyed locations and a matrix of 1's and 0's identifying if they were impacted by a storm during the previous 24 months.
3. A table with daily intersection information for the month the survey was conducted.
4. A long table with only intersection information

```{r hurricanes-main-loop}
#the website of interest for tracks
website <- 'https://www.nhc.noaa.gov/gis/archive_besttrack.php'

#lookup table with names that was created in an earlier step
storm_names_dir <- r"(./data/HurricaneRawData/storm_lookup_table.csv)"
##############################################################
#postcode files should have matching column titles to do this in a loop:
#id column
#long, lat
#survey_year, survey_month
# make sure that these column names exist
long_col <- "long"
lat_col <- "lat"
survey_id_col <- "hospital_id"
survey_year_col <- "survey_year"
survey_month_col <- "survey_month"
#############################################################
cols <- c("geometry", "event_id", "event_r_id", "storm_id")
#buffer distances in metres to go around point of interest
bufs <- c(0)#, 25000, 50000, 100000)
#years of survey dates for table naming conventions
years <- c(2023)
#this is paried with the above years, should match up!
survey_hospitals <- r"(./data/odhf_hospitals.csv)"
dirs <- c(survey_hospitals)
#results storage
hurricane_results <- list()


#this is a double loop, the first one goes through each survey year of interest 
for (k in seq_along(years)){
  message("Naming suffix: ", years[k])
  #the second loop applies the buffer size to the survey data for intersection analysis 
  # between storm pathway and survey data
  for (i in seq_along(bufs)){
    message("Point buffer extent: ", bufs[i]/1000, " km")
    #read the survey data
    survey_data <- read.csv(dirs[k])
    # convert the survey data to an sf object for the intersection
    survey_sf <- st_as_sf(
      #eliminate the Nas
      survey_data[!is.na(survey_data[[long_col]]) & !is.na(survey_data[[lat_col]]), ],
      coords = c(long_col, lat_col),
      crs = 4326
    )
    #make sure its the appropriate crs
    survey_sf <- st_transform(survey_sf, 4326)
    #apply the buffer to the survey points
    survey_buffer <- st_buffer(survey_sf, bufs[i])
    #create a date type column from the survey year and survey month columns
    survey_buffer$months <- ym(paste0(survey_buffer[[survey_year_col]], survey_buffer[[survey_month_col]]))
    survey_buffer$survey_id <- paste(survey_buffer[[survey_id_col]], survey_buffer[[survey_month_col]], survey_buffer[[survey_year_col]], sep="_")
    #get all storms that happen within 30 months of the earliest survey date and the latest survey month of that year
    # this could be a pretty variable range of dates
    all_storms_within_range <- all_storms %>% filter(months >= (min(survey_buffer$months)-months(30)) & months <= max(survey_buffer$months))
    #do the intersection between the surveys and all of those storms
    intersection <- st_intersects(survey_buffer, all_storms_within_range, sparse = TRUE)
    
    #build the dataframe
    intersection_df <- map2_dfr(
      #for each intersection
      seq_along(intersection),
      #create a little tibble of point, strom, survey, and if it intersects
      # creates an association between each poi and storm intersect event and attaches the date of the survey
      intersection,
      ~ tibble(
        point_id = survey_sf[[survey_id_col]][.x],
        storm_id = all_storms_within_range$event_r_id[.y],
        survey_date = survey_buffer$months[.x],
        intersects = 1
      )
    )
    
    #get the strom names
    storm_lookup <- read.csv(storm_names_dir, sep=",")
    
    #preserve the intersection dataframe
    matrix_df <- intersection_df
    #get the survey data anew
    survey_data <- read.csv(dirs[k])
    #set blanks to na's
    survey_data[survey_data == ""] <- NA
    #only keep non na points
    survey_data <- survey_data[!is.na(survey_data[[long_col]]) & !is.na(survey_data[[lat_col]]), ]
    #convert to data.table
    survey_data <- as.data.table(survey_data)
    #organize dates
    survey_data[, dates := ceiling_date(make_date(survey_year, survey_month, 1), "month") - days(1)]
    
    #convert to data.table
    matrix_df <- as.data.table(matrix_df)
    #only interested in the intersections, there shouldn't be any misses, but just in case
    matrix_df <- matrix_df[intersects == 1]
    #set the date column in the matrix_df data frame
    matrix_df[, year_month := format(ym(str_sub(storm_id, 5, 10)), "%Y%m")]
    #only need to know if there is an intersection on that month, 
    # so get unique entries for each poi and month combination
    matrix_df <- unique(matrix_df, by = c("point_id","survey_date", "year_month"))
    
    #expand each survey row into last 24 months (month_0 = survey month, month_1 = survey-1, etc)
    expand_months <- function(d, s_id, survey_month, survey_year) {
      data.table(
        survey_id = s_id,
        survey_year = survey_year,
        survey_month = survey_month,
        offset = 0:24,
        date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
      )
    }
    
    #apply the function
    # identifies how many months a given month is offset from the 0th month for that pois survey date
    survey_expanded <- survey_data[, expand_months(dates, get(survey_id_col), survey_month, survey_year), by = c(survey_id_col, survey_year_col, survey_month_col)]
    survey_expanded$survey_date <- ym(paste0(survey_expanded[[survey_year_col]], survey_expanded[[survey_month_col]]))

    #join the storms on survey id and the date at year_month
    result <- matrix_df[survey_expanded, 
                        on = c("point_id" = survey_id_col, "year_month" = "date")]
    
    #add 0's to the matrix of 1's and 0's
    result[is.na(intersects), intersects := 0]
    
    #join storm names for easier identification
    storm_lookup <- as.data.table(storm_lookup)
    #storm id and year to cross reference with website.
    storm_lookup[, storm_id_year := paste0(storm_id, year)]
    result[, storm_id_year := str_sub(storm_id, 1, 8)]
    result <- merge(
      result,
      storm_lookup[, .(storm_id_year, storm_name)],
      by = "storm_id_year",
      all.x = TRUE
    )
    
    #pivot to one row per id and columns month_0 to 24\
    formula_str <- paste("survey_id + survey_month + survey_year ~ paste0('month_', offset)")
    result_months <- dcast(
      result,
      as.formula(formula_str),
      value.var = "intersects",
      fun.aggregate = max
    )
    
    #join the lat and long from the original dataset back for joining later
    on_vec <- setNames(c(survey_id_col, "survey_year", "survey_month"), 
                   c("survey_id", "survey_year", "survey_month"))

    result_months[survey_data, 
                  on = on_vec,
                  `:=`(lat = i.lat, long = i.long)]
    
    ################### month 0 information day by day ###############################
    matrix_df <- intersection_df
    matrix_df <- matrix_df %>% 
      #only intersections
      filter(intersects == 1) %>%
      #get date to the day of the storm
      mutate(year_month_day = ymd(str_sub(storm_id, 5, 12))) %>% 
      mutate(storm_id_long = str_sub(storm_id, 0, 12)) %>% 
      #one row per day, doesn't matter the wind speed
      distinct(., point_id, year_month_day,.keep_all=TRUE) %>% 
      #format date as a date again
      mutate(year_month_day = format(year_month_day, "%Y-%m-%d"))
    
    #get survey data  
    survey_data <- read.csv(dirs[k])
    #set blanks to na's
    survey_data[survey_data == ""] <- NA
    #only keep non na points
    survey_data <- survey_data[!is.na(survey_data[[long_col]]) & !is.na(survey_data[[lat_col]]), ]
    #assume survey was conducted on the final day of the month, so check every day of the month
    survey_data$dates <- ceiling_date(make_date(year=survey_data[[survey_year_col]], month = survey_data[[survey_month_col]], day = 1), "month") - days(1)
    
    #expand to every day of the month
    expand_survey_data <- survey_data %>%
      rowwise() %>%
      mutate(days_in_month = list(seq(from = floor_date(dates, "month"), to = dates, by = "1 day"))) %>%
      unnest(days_in_month) %>%
      rename(date = days_in_month) %>% 
      mutate(date = format(date,"%Y-%m-%d"))
    
    #join the two datasets of storm data and survey data
    result_days <- left_join(expand_survey_data, matrix_df, by = c("date" = "year_month_day", setNames("point_id", survey_id_col)))
    
    #add to final dataframe of dataframes
    key <- paste0("result_", bufs[i]/1000, "k_", years[k])
    hurricane_results[[key]] <- list(
      long = result,
      wide = result_months,
      daily = result_days,
      matrix = matrix_df
    )
  }
}
```

# Appendix

## Survey Data Generation

```{r sample-data-generation}
#TODO incorporate the reference for this data into the index
#generation of data based on entries in the hospital data 
# if there are 20 hospitals, make 20 random months for survey between 1 and 12 and random years choosing between 2016 and 2020
# https://www150.statcan.gc.ca/n1/en/pub/13-26-0001/2020001/ODHF_v1.1.zip
# make sure the latitude and longitude columns exist and that there is an id column as well
#set a random seed
set.seed(933)
#read in the raw data
odhf <- read.csv("./data/odhf_v1.1.csv")
#create an id list
odhf$hospital_id <- 1:nrow(odhf)
#can only use valid lat/long locations
odhf_valid <- odhf[!is.na(odhf[["latitude"]]) & 
                             !is.na(odhf[["longitude"]]), ]
odhf_valid <- odhf_valid %>% filter(odhf_facility_type=="Hospitals") %>%  select(c("latitude", "longitude", "facility_name", "odhf_facility_type", "hospital_id"))
odhf_valid[[survey_month_col]] <- sample(1:12, size = nrow(odhf_valid), replace = TRUE)
odhf_valid[[survey_year_col]] <- sample(2018:2021, size=nrow(odhf_valid), replace = TRUE)
setnames(odhf_valid, c("latitude", "longitude"), c("lat","long"))
write.csv(odhf_valid, "./data/odhf_hospitals.csv")
```

