# Locating & Refining Environmental Data {#locating-and-refining-environmental-data}

## Introduction {#introduction-environmental-data}

Sample data is the hospital data.

## Wildfires {#wildfires}

### Data preparation with sf package {#data-preparation-with-sf-package}

We read the updated NBAC wildfire polygons (1972-2024) from the .shp file.

-   Tip: Shapefiles are a set of files. Ensure the .shp, .shx, .dbf, and (ideally) .prj sit together in the same folder.

```{r}
poly_fires <- st_read(r"(../data/FireData/NBAC_1972to2024_20250506.shp)") 

names(poly_fires)
```

### Inspect & refine data structure {#inspect-and-refine-data-structure}

Check available columns and data types

```{r}
names(poly_fires)
str(poly_fires)
```

Normalise column names

```{r}
names(poly_fires) <- tolower(names(poly_fires)) 
names(poly_fires)
```

Given the extremely large size, it's best to reduce the dataset to include only the years you require

```{r}
poly_fires <- poly_fires %>% 
  filter(year >= 2010)
```

Limit dataset to only the columns you require

```{r}
poly_fires <- select(poly_fires, 
                     nfireid, basrc,
                     year, 
                     ag_sdate, # start date as date format
                     ag_edate, # end date as date format
                     poly_ha, 
                     admin_area,
                     geometry)

head(poly_fires)
```

Rename columns to align better with your spatial datasets

```{r}
poly_fires <- poly_fires %>%
  mutate(
    start_day = day(ag_sdate),
    start_month = month(ag_sdate),
    start_year = year(ag_sdate),
    end_day = day(ag_edate),
    end_month = month(ag_edate),
    end_year = year(ag_edate),
    start_date = ag_sdate, 
    end_date = ag_edate,
    size = poly_ha) %>%
  select(-poly_ha)  # remove the original poly_ha column


names(poly_fires)
str(poly_fires)
```

You may also want to save this newly formed dataset for easier future loading (i.e., to avoid the need to reload the shapefile during your next R session)

```{r}
# uncomment to run
# write.csv(poly_fires, "wildfire_data.csv", row.names = FALSE)
```

## Hurricanes {#hurricanes}

### Data Sourcing {#hurricanes-data-sourcing}

Hurricane data was sourced from the National Oceanic and Atmospheric Administration (NOAA). Data can be obtained programmatically (as below), or manually for each hurricane of interest from the [NOAA website](https://www.nhc.noaa.gov/gis/archive_besttrack.php?year= "Hurricane Best Tracks"). Should you choose to download the files yourself, the script is designed to use shapefiles (.shp).

-   It is important to note that the following script should be used sparingly

We first need to specify where we are getting the hurricane data from and for which years.

```{r}
base_page <- "https://www.nhc.noaa.gov/gis/archive_besttrack.php?year="
base_zip  <- "https://www.nhc.noaa.gov/gis/best_track/"
hurricane_years_interest <- 2008:2025 #getting data for all years from 2008 to 2025
```

Next, we set up the storage location for the hurricane best track files and create a lookup table for later. It is important to set a viable output directory for the best tracks from the website.

```{r}
#set an output directory for the data to download into
base_shape_dir <- "../data/HurricaneRawData/nhc_best_tracks"
dir.create(base_shape_dir, showWarnings = FALSE)

#create a lookup table for the storms that can be used later to create a connection between the storm name and data
storm_summary <- data.frame(Year = integer(), Storms = integer())
storm_lookup  <- data.frame(year = integer(), storm_id = character(), storm_name = character())
```

Finally, get the files from the website for each year specified. Within this snippet, a lookup table for the file names and actual hurricane name is created and saved as a .csv for later reference. (TODO: add a switch to get just the names list vs. the file and names list)

```{r, eval = FALSE}
#process every year for the year range specified above
for (year in hurricane_years_interest) {
  cat("processing", year, "\n")
  
  #read the page, catch if the response is blank
  page <- tryCatch({
    read_html(paste0(base_page, year))
  }, error = function(e) {
    cat("could not read page for", year, "\n")
    return(NULL)
  })
  if (is.null(page)) next
  
  #extract all the rows on the page to get the hurricane codes and names from the table for that year
  rows <- html_nodes(page, "tr")
  
  #loop through rows to match "al##" and extract name from second column to create lookup table
  for (row in rows) {
    cols <- html_nodes(row, "td") #get all the td html elements
    
    if (length(cols) >= 2) {
      id_text <- str_trim(html_text(cols[1])) #just want the id from the td elements
      
      if (str_detect(tolower(id_text), "^al\\d{2}$")) { #regex to get the atlantic (al) storm and the id number
        name_text <- str_trim(html_text(cols[2])) #get the nominal identifier
        storm_lookup <- rbind( #collect and add to the storm loookup table
          storm_lookup,
          data.frame( #create a data frame with the year, storm id from the td element, and the storm name
            year = year,
            storm_id = tolower(id_text),
            storm_name = name_text,
            stringsAsFactors = FALSE
          )
        )
      }
    }
  }
  
  #get just the text from all the td html elements
  td_text <- html_text(html_nodes(page, "td"), trim = TRUE)
  
  #get the strom ids again (for redundancies sake)
  storm_ids <- unique(str_to_lower(str_subset(td_text, "^al\\d{2}")))
  
  #make sure it's just the 4 character id with more regex
  storm_ids <- gsub("\\s.*", "", storm_ids)
  
  cat(" found", length(storm_ids), "atlantic storms\n")
  
  #storm summary for creating storm ids in file names
  storm_summary <- rbind(storm_summary,
                         data.frame(Year = year, Storms = length(storm_ids)))
  
  #file naming system is regular within the website hierarchy, so download all zip shapefiles using the engineered storm names
  for (storm_id in storm_ids) {
    zip_name <- paste0(storm_id, year, "_best_track.zip")
    url <- paste0(base_zip, zip_name)
    dest <- file.path(base_shape_dir, zip_name) #set outpath based on string specified earlier
    
    cat(" downloading:", zip_name, "\n")
    resp <- tryCatch({
      GET(url, write_disk(dest, overwrite = TRUE)) #get zip file from constructed url
    }, error = function(e) {
      cat(" failed to download", zip_name, "\n") #sometimes there are no files to download, especially in earlier years
      return(NULL)
    })
    
    #unzip the files too, easier to do now than later
    if (!is.null(resp) && status_code(resp) == 200) {
      unzip(dest, exdir = file.path(base_shape_dir, paste0(storm_id, "_", year)))
    } else {
      cat(" file not found:", url, "\n")
    }
    #be kind to the server and don't get banned from scraping
    Sys.sleep(1.1)
  }
}

write.csv(storm_lookup, r"(../data/HurricaneRawData/storm_lookup_table.csv)")
```

### Data Cleaning {#hurricanes-data-cleaning}

Now that the data is downloaded and ready, it is time clean it up. There are two files of interest for the hurricanes, the point files and the polygon (radii) files. These files need to be manipulated so that dates and ids are consistent between the two file sets and between years. 

```{r}
#standardizing naming conventions in files
month_map <- c(
  "jan" = 1, "january" = 1, "01" = 1,
  "feb" = 2, "february" = 2, "02" = 2,
  "mar" = 3, "march" = 3, "03" = 3,
  "apr" = 4, "april" = 4, "04" = 4,
  "may" = 5, "05" = 5,
  "jun" = 6, "june" = 6, "06" = 6,
  "jul" = 7, "july" = 7, "07" = 7,
  "aug" = 8, "august" = 8, "08" = 8,
  "sep" = 9, "september" = 9, "09" = 9,
  "oct" = 10, "october" = 10, "10" = 10,
  "nov" = 11, "november" = 11, "11" = 11,
  "dec" = 12, "december" = 12, "12" = 12
)

#base directory where files are saved
base_dir <- r"(../data/HurricaneRawData/nhc_best_tracks)"

#set an ID for all of the files in the pts files and radii files
#get all shapefiles recursively
shp_files <- list.files(base_dir, pattern = "(_pts\\.shp$|_radii\\.shp$)", recursive = TRUE, full.names = TRUE)
```

Add the storm ID for each storm to the file names and overwrite the shapefile.

```{r}
#process the files
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  #read the shapefile
  shp <- st_read(shp_path, quiet = TRUE)
  
  #extract part before first underscore
  file_prefix <- sub("_.*", "", basename(shp_path))
  
  #add the storm id file
  shp <- shp %>% mutate(storm_id = file_prefix)
  
  #overwrite shapefile
  st_write(shp, shp_path, delete_dsn = TRUE, quiet = TRUE)
}
```


Date and time need to be consistent between the point and polygon files.

```{r}
#Create a new directory for processed files
output_dir <- file.path(base_dir, "processed")

#processing for just the _pts.shp files to normalize dates and times
base_dir <- r"(../data/HurricaneRawData/nhc_best_tracks)"
output_dir <- file.path(base_dir, "processed")
dir.create(output_dir, showWarnings = FALSE)

#get all the shapefiles recursively
shp_files <- list.files(base_dir, pattern = "_pts\\.shp$", recursive = TRUE, full.names = TRUE)

#recursively alter the pts files to the normalized date and time fields with ids
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  shp <- st_read(shp_path, quiet = TRUE)
  
  #files had different formats for dates and times, this normalizes them
  # important step to combine everything for later analysis
  shp <- shp %>%
    mutate(
      mm = str_pad(as.integer(unname(month_map[tolower(MONTH)])), 2, pad = "0"),
      dd = str_pad(as.integer(DAY), 2, pad = "0"),
      hh = str_sub(HHMM, 1, 2),
      event_id = tolower(paste0(storm_id, mm, dd, hh))
    )
  
  #preserve folder structure in output
  rel_path <- path_rel(shp_path, start = base_dir)
  out_path <- file.path(output_dir, rel_path)
  dir_create(dirname(out_path))
  
  #write output
  st_write(shp, out_path, delete_dsn = TRUE, quiet = TRUE)
}
```


The points files also need to be done in a similar way.

```{r}
#slightly different process for the _radii file they already have the synoptime field and those just need to be modified
shp_files <- list.files(base_dir, pattern = "_radii\\.shp$", recursive = TRUE, full.names = TRUE)

#recursively alter the pts files to fie normalized date and time fields with ids
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  shp <- st_read(shp_path, quiet = TRUE)
  
  #create matching event_id to the points fields so they can be related (since they have different kinds of information)
  shp <- shp %>%
    mutate(
      event_id = tolower(
        paste0(storm_id, substr(SYNOPTIME, 5, nchar(SYNOPTIME))) #this will be storm id, date, and UTC hour in 24 hour format i.e. [al01][20250831][12]
      )
    )
  
  #different kind of id, radii refers to the range of wind at that knot level i.e. _34 would be 34 knot wind
  shp <- shp %>%
    mutate(
      event_r_id = tolower(
        paste0(storm_id, substr(SYNOPTIME, 5, nchar(SYNOPTIME)), "_", RADII)#this will be storm id, date, and UTC hour in 24 hour format, and radii distance  i.e. [al01][20250831][12][_34]
      )
    )
  
  #preserve folder structure
  rel_path <- path_rel(shp_path, start = base_dir)
  out_path <- file.path(output_dir, rel_path)
  dir_create(dirname(out_path))
  
  #write output
  st_write(shp, out_path, delete_dsn = TRUE, quiet = TRUE)
}
```


### Data Manipulation {#hurricanes-data-manipulation}

Now that the hurricane files are prepared properly, the associated points data of interest, (in this case hospitals), can be used within the files. THe point files in this example have a series of months and years in columns when surveys were conducted, a latitude and longitude column where we can identify

First we define the functions to manipulate the data and set a configuration.

```{r}

Sys.setenv("OGR_GEOMETRY_ACCEPT_UNCLOSED_RING" = "YES")
CONFIG <- list(
  website = 'https://www.nhc.noaa.gov/gis/archive_besttrack.php',
  storms_dir = r"(../data/HurricaneRawData/nhc_best_tracks/processed)",
  storm_names_dir = r"(../data/HurricaneRawData/storm_lookup_table.csv)",
  coord_cols = list(long = "long", lat = "lat"),
  storm_cols = c("geometry", "event_id", "event_r_id", "storm_id"),
  buffer_distances = c(0), #in metres
  crs = 4326,
  lookback_months = 30,
  history_months = 24
)

#function to load the storm data from user specified directories
load_storm_data <- function(storms_dir, cols, crs = CONFIG$crs) {
  #get all the radii files, since these are the polygon files
  shp_files <- list.files(storms_dir, pattern = "_radii\\.shp$", 
                          recursive = TRUE, full.names = TRUE)
  
  #load each shapefile separately to avoid corruption
  storm_list <- lapply(shp_files, function(shp) {
    st_read(shp, quiet = TRUE) %>% 
      select(any_of(cols)) %>%
      st_transform(crs) %>%
      #this throws OGR warnings for unclosed fans
      st_make_valid()
  })
  
  # Combine after all are processed
  all_storms <- do.call(rbind, storm_list)
  #extract part of the date
  all_storms$months <- ym(str_sub(all_storms$event_id, 5, 10))
  
  all_storms
}

#load the survey point data from a csv, set as a generic crs unless otherwise specified
load_survey_data <- function(file_path, long_col, lat_col, crs = CONFIG$crs) {
  survey_data <- read.csv(file_path)
  survey_data[survey_data == ""] <- NA
  
  valid_data <- survey_data[!is.na(survey_data[[long_col]]) & 
                             !is.na(survey_data[[lat_col]]), ]
  
  #convert survey data to an sf object to compute intersections
  survey_sf <- st_as_sf(
    valid_data,
    coords = c(long_col, lat_col),
    crs = crs
  ) %>% st_transform(crs)
  
  list(data = valid_data, sf = survey_sf)
}

#the max and min dates are integers, not date objects
#TODO handle any date that is passed, extracting the months 
filter_storms_by_date_range <- function(storms, min_date, max_date, lookback_months) {
  storms %>% 
    filter(months >= (min_date - months(lookback_months)) & 
           months <= max_date)
}

#create a buffer around the survey data if looking for incidence within a radius
prepare_survey_buffer <- function(survey_sf, buffer_dist) {
  #extract attributes before buffering to errors
  id <- survey_sf$id
  survey_year <- survey_sf$survey_year
  survey_month <- survey_sf$survey_month
  
  #buffer the survey data
  survey_buffer <- st_buffer(survey_sf, buffer_dist)
  
  #reattach the appropriate attributes
  survey_buffer$id <- id
  survey_buffer$survey_year <- survey_year
  survey_buffer$survey_month <- survey_month
  survey_buffer$months <- ym(paste0(survey_year, survey_month))
  survey_buffer$survey_id <- paste(id, survey_month, 
                                          survey_year, sep = "_")
  survey_buffer
}


#get the intersections between the buffered survey data and the storms
compute_storm_intersections <- function(survey_buffer, storms) {
  #extract attributes to avoid type 29 error
  ids <- survey_buffer$id
  survey_dates <- survey_buffer$months
  storm_ids <- storms$event_r_id
  
  #get intersections between survey infomration and storms
  intersection <- st_intersects(survey_buffer, storms, sparse = TRUE)
  
  #there was an error unless the results list was defined like this for the next loop
  result_list <- vector("list", length(intersection))
  
  #loop through the intersections and add the information to the list
  for (i in seq_along(intersection)) {
    if (length(intersection[[i]]) > 0) {
      result_list[[i]] <- data.frame(
        point_id = ids[i],
        storm_id = storm_ids[intersection[[i]]],
        survey_date = survey_dates[i],
        intersects = 1,
        stringsAsFactors = FALSE
      )
    }
  }
  
  bind_rows(result_list)
}

#want to know occurrences going back 'x' amount of months in the past
expand_survey_months <- function(survey_data, history_months) {
  survey_dt <- as.data.table(survey_data)
  #surveys not conducted on the same days or sometimes over multiple days so use the last day of every month as the limit for the surveys 
  survey_dt[, dates := ceiling_date(make_date(survey_year, survey_month, 1), 
                                    "month") - days(1)]
  
  #standardizing the expansion of months into the past
  expand_months <- function(d, id, survey_month, survey_year) {
    data.table(
      id = id,
      survey_year = survey_year,
      survey_month = survey_month,
      offset = 0:history_months,
      date = format(ceiling_date(d %m-% months(0:history_months), "month") - 
                    days(1), "%Y%m")
    )
  }
  
  #when the expansion through months occurs, keep unique entries based on id_number, survey year, and survey month
  survey_dt[, expand_months(dates, id, survey_month, survey_year), 
            by = .(id, survey_year, survey_month)]
}

#intersection information down to the day for the month of intersection
expand_survey_days <- function(survey_data) {
  #assume the survey data was collected at the end of the month
  survey_data$dates <- ceiling_date(
    make_date(year = survey_data$survey_year, 
              month = survey_data$survey_month, 
              day = 1), 
    "month"
  ) - days(1)
  
  #get the day of intersection for month 0 (the month where intersections happen)
  survey_data %>%
    rowwise() %>%
    mutate(days_in_month = list(seq(from = floor_date(dates, "month"), 
                                    to = dates, by = "1 day"))) %>%
    unnest(days_in_month) %>%
    rename(date = days_in_month) %>% 
    mutate(date = format(date, "%Y-%m-%d"))
}

#process the results on a monthly basis
process_monthly_results <- function(intersection_df, survey_data, 
                                   storm_lookup_path, history_months) {
  
  #put the intersection data into a data.table and isolate only the entries that intersect
  matrix_dt <- as.data.table(intersection_df)
  matrix_dt <- matrix_dt[intersects == 1]
  #the month and year that the survey happened in should be two different columns, make that into a date
  matrix_dt[, year_month := format(ym(str_sub(storm_id, 5, 10)), "%Y%m")]
  #get only unique points based on the survey date and id
  matrix_dt <- unique(matrix_dt, by = c("point_id", "survey_date", "year_month"))
  
  #expand the survey data 'x' number of months into the past 
  survey_expanded <- expand_survey_months(survey_data, history_months)
  survey_expanded$survey_date <- ym(paste0(survey_expanded$survey_year, 
                                           survey_expanded$survey_month))
  
  #merge the results and filter for only intersections again
  result <- matrix_dt[survey_expanded, 
                      on = .(point_id = id, year_month = date)]
  result[is.na(intersects), intersects := 0]
  
  #add storm names for easier reading
  storm_lookup <- fread(storm_lookup_path)
  storm_lookup[, storm_id_year := paste0(storm_id, year)]
  result[, storm_id_year := str_sub(storm_id, 1, 8)]
  result <- merge(result, storm_lookup[, .(storm_id_year, storm_name)],
                  by = "storm_id_year", all.x = TRUE)
  
  #pivot to wide format
  result_wide <- dcast(
    result,
    id + survey_month + survey_year ~ paste0("month_", offset),
    value.var = "intersects",
    fun.aggregate = max
  )
  
  #add spatial information back to the data
  survey_dt <- as.data.table(survey_data)
  result_wide <- result_wide[survey_dt, 
                             on = .(id, survey_year, survey_month),
                             `:=`(lat = i.lat, long = i.long)]
  
  list(long = result, wide = result_wide)
}

#get month 0 daily results to determine exact date of hurricane intersection
process_daily_results <- function(intersection_df, survey_data) {
  #get distinct days and ensure proper date and storm id formatting
  matrix_daily <- intersection_df %>% 
    filter(intersects == 1) %>%
    mutate(year_month_day = ymd(str_sub(storm_id, 5, 12)),
           storm_id_long = str_sub(storm_id, 0, 12)) %>% 
    distinct(point_id, year_month_day, .keep_all = TRUE) %>% 
    mutate(year_month_day = format(year_month_day, "%Y-%m-%d"))
  
  #expand to all days in a month
  expanded_days <- expand_survey_days(survey_data)
  
  left_join(expanded_days, matrix_daily, 
            by = c("date" = "year_month_day", "id" = "point_id"))
}

```

Load all the storm data from the files stored locally.

```{r}

#load the storm data
#TODO fix the warnings, can safely supress them
message("Loading storm data...")
all_storms <- load_storm_data(CONFIG$storms_dir, CONFIG$storm_cols, CONFIG$crs)
```

This function ties all the defined functions together into one, processing each survey data file of interest with each of the hurricane data files of interest. 

```{r}
analyze_hurricanes <- function(survey_dirs, all_storms, years, config = CONFIG) {
  
  #define a list of results if multiple surveys are being used
  results <- list()
  
  #outer loop for years of interest for saving the files
  for (k in seq_along(years)) {
    message("Processing year: ", years[k])
    
    #can set multiple buffer distances in the config, leave at 0 for a simple intersection
    for (i in seq_along(config$buffer_distances)) {
      buf <- config$buffer_distances[i]
      message("  Buffer: ", buf / 1000, " km")
      
      #load the survey data for this iteration
      survey <- load_survey_data(survey_dirs[k], 
                                 config$coord_cols$long, 
                                 config$coord_cols$lat, 
                                 config$crs)
      
      #force garbage collection for larger processing times
      gc()
      
      #prepare the survey buffer
      survey_buffer <- prepare_survey_buffer(survey$sf, buf)
      
      #filter storms by date range
      storms_in_range <- filter_storms_by_date_range(
        all_storms, 
        min(survey_buffer$months), 
        max(survey_buffer$months),
        config$lookback_months
      )
      
      message("    Determining intersections...")
      intersection_df <- compute_storm_intersections(survey_buffer, storms_in_range)
      
      #drop geometry columns to avoid type 29 errors
      survey_data_clean <- survey$data
      
      message("    Processing monthly results...")
      monthly <- process_monthly_results(
        intersection_df, 
        survey_data_clean, 
        config$storm_names_dir,
        config$history_months
      )
      
      message("    Processing daily results...")
      daily <- process_daily_results(intersection_df, survey_data_clean)
      
      #store results in nested lists
      key <- paste0("result_", buf / 1000, "k_", years[k])
      results[[key]] <- list(
        long = monthly$long,
        wide = monthly$wide,
        daily = daily,
        matrix = intersection_df
      )
      
      #clean up spatial objects
      rm(survey_buffer, storms_in_range)
      gc()
    }
  }
  
  results
}
```


Sample run with some sample data (public hospital information). The survey data was generated randomly using the code in the appendix. 

```{r}

#2023
survey_dir_2023 <- r"(../data/)"

# Usage Example ----
# Define your survey directories and years
survey_dirs <- c(survey_dir_2023, survey_dir_2018, survey_dir_2014)
years <- c(2023, 2018, 2014)

hurricane_results <- analyze_hurricanes(survey_dirs,all_storms, years)
```

#Appendix

##Survey Data Generation

```{r}
#generation of data based on entries in the hospital data 
# if there are 20 hospitals, make 20 random months for survey between 1 and 12 and random years choosing between 2016 and 2020
# make sure the latitude and longitude columns exist and that there is an id column as well
```

