```{r setup-heat, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/matth/Climate_and_Health/bookdown_project')
```

# Heat {#heat}

Huge swings in temperature are becoming more and more common when comparing seasonal highs and lows. It is much harder to stay cool in heat...

## Environmental Data Preparation {#heat-data-preparation}

Our first step is to load and transform the hospital data. Temperature data was sourced from the `weathercan` R package. 

```{r load-heat, message=FALSE, results='hide'}
#load the data
odhf_valid <- read.csv("./data/odhf_hospitals.csv", row.names = 1)
survey_sf <- odhf_valid %>% st_as_sf(coords=c("long", "lat"), crs=st_crs(4326)) %>% select(-facility_name)
survey_sf <- st_transform(survey_sf, 3978)
stations_list <- stations() 
names(stations_list)
```

In order to collect the relevant weather stations and time frames, the minimum and maximum survey data collection periods needed to be collected from the ODHF data.

```{r set-weather-range, message=FALSE}
#identify the date range
survey_sf <- survey_sf %>%
  mutate(
    weather_start = make_date(survey_year - 2, survey_month, 1),  #Start date: 2 years before
    weather_end = make_date(survey_year, survey_month, days_in_month(make_date(survey_year, survey_month, 1))))

min_start_year <- min(year(survey_sf$weather_start), na.rm = TRUE)
max_end_year <- max(year(survey_sf$weather_end), na.rm = TRUE)

stations_by_day <- filter(stations_list, 
                                      interval %in% c("hour","day"), 
                                      #setting the min coverage years for data at the stations
                                      #is to increase likelihood that the same hospital location will be assigned                                          to the same weather station
                                      #across survey years 
                                      #this will reduce noise introduced as a result of pulling data for the same                                         location from different sources 
                                      #which may have different data available, different frequencies of                                                 collection, etc. 
                                      start <= min_start_year, 
                                      end >= max_end_year)
```

Next, we need to get the nearest distance to stations for evaluation of if they are too close or too far away. In the case of this example, no station will be too far away.

```{r station-filtering}
#get the stations by day
stations_by_day_sf <- st_as_sf(stations_by_day, coords=c("lon", "lat"), crs = 4326)
stations_by_day_sf <- st_transform(stations_by_day_sf, st_crs(survey_sf))
#get the nearest feature
nearest_stations <- st_nearest_feature(survey_sf, stations_by_day_sf)
#get the nearest distances and ids of the stations for each hospital
survey_sf$nearest_dist <- st_distance(survey_sf, stations_by_day_sf[nearest_stations,], by_element = TRUE)
survey_sf$near_station_id <- stations_by_day_sf[nearest_stations,]$station_id
survey_sf$station_interval <- stations_by_day_sf[nearest_stations,]$interval
survey_sf <- survey_sf %>% rename(station_id=near_station_id)
```

Here, we get the data from the `weathercan` package.

```{r get-weather-data, eval=FALSE}
#query the package for every station that was identified
hourly_station_ids <- filter(survey_sf, station_interval == "hour") %>% pull(station_id)
daily_station_ids <- filter(survey_sf, station_interval == "day") %>% pull(station_id)
daily_weather <- weather_dl(station_ids = daily_station_ids, start=min(survey_sf$weather_start), end=max(survey_sf$weather_end), interval='day')
hourly_weather <- weather_dl(station_ids = hourly_station_ids, start=min(survey_sf$weather_start), end=max(survey_sf$weather_end), interval='hour')
```

Then, we group and combine the hourly with the daily so a comprehensive dataset is created.

```{r group-weather-data, eval=FALSE}
#summarize from hourly to daily and combine
hourly_as_daily_weather_original <- hourly_weather %>% 
  group_by(station_id, date, lat, lon) %>% 
  summarise(
    min_temp = min(temp, na.rm = FALSE),
    max_temp = max(temp, na.rm = FALSE),
    .groups = "drop"
  )
hourly_as_daily_weather <- subset(hourly_as_daily_weather_original, select = -c(lat, lon))
combined_hourly_daily <- bind_rows(hourly_as_daily_weather, daily_weather) %>%
  select(any_of(intersect(names(hourly_as_daily_weather), names(daily_weather))))
write_parquet(combined_hourly_daily, r"(./data/odhf_weather_stations.parquet)")
```

## Heat ECCC {#heat-daily-eccc}

Environment and Climate Change Canada has their own criteria for extreme heat events around the country. Using a rough sketch of the [criteria map](http://www.canada.ca/en/environment-climate-change/services/environmental-indicators/extreme-heat-events.html), we can relate the hospital data to the temperature data.

```{r heat-combined-daily-eccc, results='hide', warning=FALSE, message=FALSE}
combined_hourly_daily <- read_parquet(r"(./data/odhf_weather_stations.parquet)")
eccc_friendly <- st_read(r"(./data/roughEcccMap_final.gpkg)")
survey_sf_eccc <- st_join(survey_sf, eccc_friendly)

consecutive_days <- 2 #2 days and nights as the criteria.


#identify days of consecutive heat
heat_flagged_eccc <- survey_sf_eccc %>%
  left_join(combined_hourly_daily, by='station_id') %>%  #date, station_id, max_temp, min_temp 
  unique(.,by=c("hospital_id", "survey_month", "survey_year", "date")) %>%  #get unique entries based on survey_id, survey month + year, and date of station data collection
  arrange(station_id, date) %>% #need dates in order chronologically
  group_by(station_id, hospital_id, survey_month, survey_year) %>% #station_ids grouped together
  mutate(
    hot = max_temp >= day_thresh & min_temp >= nite_thres, #if the max_temp is greater than the day threshold AND the minimum temperature is greater than the nighttime threshold
    run_id = rleid(hot) #add unique id for runs
  ) %>% 
  group_by(hospital_id, survey_month, survey_year, station_id, run_id) %>% #group by station id and the run ids (looking for two or more)
  mutate(
    run_len = sum(hot), #sum the number of hot events per run                                    
    heat_event = hot & run_len >= consecutive_days #only classed as a heat event if two or more days       
  ) %>%
  ungroup() %>%
  select(-run_id, -run_len) %>% 
  mutate(date = as.Date(date))
```

### Date Lag {#heat-eccc-date-lag}

As part of the exposure determination, we are interested in the 24 months prior to each survey date at the hospitals. Hospitals were surveyed at different times, so the date of the survey will be recorded along with each month from the 0th to 24th month prior (survey month, survey month - 1, survey month - 2, etc.).

```{r heat-eccc-date-lag, results='hide'}
expand_months <- function(d, hospital_id, survey_month, survey_year, station_id) {
  data.table(
    hospital_id = hospital_id,
    survey_year = survey_year,
    survey_month = survey_month,
    station_id = station_id,
    offset = 0:24,
    date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
  )
}

#apply the function
#identifies how many months a given month is offset from the 0th month for that surveys survey date
survey_sf$dates <- ymd(paste(survey_sf$survey_year, survey_sf$survey_month, "01", sep = "-")) + months(1) - days(1)
#FOR EVERY UNIQUE survey AND SURVEY DATE COMBINATION <- this is key
surveys_un <- unique(survey_sf, by = c("survey_id","dates"))
surveys_un <- as.data.table(surveys_un)
survey_expanded <- surveys_un[, expand_months(dates, hospital_id, survey_month, survey_year, station_id), by = .(hospital_id, dates)]
survey_expanded <- survey_expanded %>% select(-3)
survey_expanded_heat_eccc <- survey_expanded %>% rename(year_month = date)
```

### Date Join {#heat-eccc-date-join}

Create heat event data for all months before narrowing the range and then join with the expanded days.

```{r heat-eccc-date-join}
#arrange the data chronologically once again and setup columns for joining, add another heat classifier since this is being grouped to months (month has_heat)
heat_monthly_eccc <- heat_flagged_eccc %>%
  st_drop_geometry() %>% #drop the geometry for faster speeds
  arrange(station_id, date) %>%
  mutate(year_month = as.integer(format(date, "%Y%m"))) %>%
  group_by(hospital_id, year_month) %>%
  summarise(
    has_heat = any(heat_event, na.rm = TRUE),                     
    .groups = "drop"
  ) %>%
  mutate(
    year  = as.integer(year_month / 100), #need a year eventually
    month = as.integer(year_month %% 100), #need a month eventually
    year_month = as.character(year_month)
  )
  

#join with the survey data
survey_with_heat_eccc <- survey_expanded_heat_eccc %>% #select(-5,-6) %>% 
  left_join(heat_monthly_eccc, 
            by = c("hospital_id", "year_month")) %>%
  #rename(year_month=dates) %>% 
  mutate(
    heat_flag = as.integer(coalesce(has_heat, FALSE))) #integer based has_heat           

heat_month_zero_daily_eccc <- filter(survey_with_heat_eccc, survey_year==year & survey_month==month) %>% st_drop_geometry()

```

### Matrix {#heat-matrix-eccc}

`survey_with_heat_eccc` has all the interesting information in it about the matrix of heat. 

```{r heat-matrix-eccc}
#create the summary matrix of 1's and 0's for the months
heat_summary_matrix_eccc <- survey_with_heat_eccc %>% 
  filter(heat_flag == 1) %>% #only the months with floods in them
  group_by(hospital_id, survey_year, survey_month, offset) %>% #group by the offsets
  slice_head(n=1) %>%
  right_join(survey_expanded_heat_eccc, by=c("hospital_id", "survey_year", "survey_month", "offset")) %>%
  mutate(heat_flag = replace_na(heat_flag, 0)) %>% #make sure there are no NAs
  ungroup() %>%
  as.data.table() %>%
  left_join(survey_sf %>% select(hospital_id, station_id, nearest_dist), by='hospital_id') %>%
  dcast(.,
      hospital_id + survey_month + survey_year + station_id + nearest_dist~ paste0("month_", offset),
      value.var = "heat_flag",
      fun.aggregate = max
    ) %>%
  setcolorder(c("hospital_id", "survey_month", "survey_year", 'station_id', 'nearest_dist', paste0("month_", 0:24))) 

write.csv(heat_summary_matrix_eccc, r"(./data/heat_matrix_eccc.csv)")
```

Display a snippet of the ECCC summary matrix.

```{r heat-eccc-matrix}
#display a snippet
kable(head(heat_summary_matrix_eccc),'html') %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(height = "300px")
```

## Heat Days {#heat-daily-arb}

You can set any threshold temperature as an arbitrary 

```{r heat-combined-daily-arb, results='hide', message=FALSE}
combined_hourly_daily <- read_parquet(r"(./data/odhf_weather_stations.parquet)")
temp_threshold <- 30
consecutive_days <- 3

#identify days of consecutive heat
heat_flagged <- combined_hourly_daily %>%
  arrange(station_id, date) %>% #arrange by station and chronologically
  unique(.,by=c("station_id", "date")) %>% #filter by only unqiue stations and dates, can't have any duplicates
  group_by(station_id) %>% #applying only to stations
  mutate(
    hot = max_temp >= temp_threshold, #it is hot if it is greater than the arbitrary threshold               
    run_id = rleid(hot) #group chronological runs together
  ) %>%
  group_by(station_id, run_id) %>% #group the runs together
  mutate(
    run_len = sum(hot), #sum the number of hot events per run 
    heat_event = hot & run_len >= consecutive_days    
  ) %>%
  ungroup() %>%
  select(-run_id, -run_len) %>% 
  mutate(date = as.Date(date))

heat_flagged_arb <- heat_flagged
```

### Date Lag {#heat-arb-date-lag}

Run the same date lags as above.

```{r heat-arb-date-lag, results='hide', message=FALSE}
expand_months <- function(d, hospital_id, survey_month, survey_year, station_id) {
  data.table(
    hospital_id = hospital_id,
    survey_year = survey_year,
    survey_month = survey_month,
    station_id = station_id,
    offset = 0:24,
    date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
  )
}

#apply the function
#identifies how many months a given month is offset from the 0th month for that surveys survey date
survey_sf$dates <- ymd(paste(survey_sf$survey_year, survey_sf$survey_month, "01", sep = "-")) + months(1) - days(1)
#FOR EVERY UNIQUE SURVEY DATE COMBINATION <- this is key
surveys_un <- unique(survey_sf, by = c("survey_id","dates"))
surveys_un <- as.data.table(surveys_un)
survey_expanded <- surveys_un[, expand_months(dates, hospital_id, survey_month, survey_year, station_id), by = .(hospital_id, dates)]
survey_expanded <- survey_expanded %>% select(-3)
survey_expanded_heat_days <- survey_expanded %>% rename(year_month = date)
```

### Date Join {#heat-arb-date-join}

Create heat event data for all months before narrowing the range and then join with the expanded days.

```{r heat-arb-date-join, results='hide', message=FALSE, warning=FALSE}
#create the monthly heat summary
heat_monthly <- heat_flagged %>%
  arrange(station_id, date) %>% #arrange chronologically
  mutate(year_month = as.integer(format(date, "%Y%m"))) %>%        
  group_by(station_id, year_month) %>% #group by year_month and station
  summarise(
    has_heat = any(heat_event, na.rm = TRUE), #we only care about any heat event in the month
    .groups = "drop"
  ) %>%
  mutate(
    year  = as.integer(year_month / 100), #need a year eventually
    month = as.integer(year_month %% 100), #need a month eventually
    year_month = as.character(year_month)
  )

survey_with_heat_days <- survey_expanded_heat_days %>% #join with monthly data
  left_join(heat_monthly, 
            by = c("station_id", "year_month")) %>%          
  mutate(
    heat_flag = as.integer(coalesce(has_heat, FALSE))) #coalesce to an integer
  
heat_flagged_daily_normalized <- heat_flagged %>% #normalized is a misnomer, we are just looking for the month 0 stuff with this one
  mutate(year_month = as.integer(format(date, "%Y%m"))) %>% 
  mutate(
    year  = as.integer(year_month / 100),      
    month = as.integer(year_month %% 100),
    year_month = as.character(year_month)
  ) %>% left_join(survey_expanded_heat_days, by = c("station_id","year_month")) %>% 
  filter(offset==0, heat_event == TRUE)
```

### Matrix {#heat-arb-matrix}

Create the summary matrix of 1's and 0's.

```{r heat-arb-matrix, message=FALSE, warning=FALSE, results='hide'}
#create the summary matrix of 1's and 0's for the months
heat_summary_matrix_arb <- survey_with_heat_days %>% 
  filter(heat_flag == 1) %>% #only the months with heat in them
  group_by(hospital_id, survey_year, survey_month, offset) %>% #group by the offsets
  slice_head(n=1) %>%
  right_join(survey_expanded_heat_days, by=c("hospital_id", "survey_year", "survey_month", "offset")) %>%
  mutate(heat_flag = replace_na(heat_flag, 0)) %>% #make sure there are no NAs
  ungroup() %>%
  as.data.table() %>%
  left_join(survey_sf %>% select(hospital_id, station_id, nearest_dist), by='hospital_id') %>%
  dcast(.,
      hospital_id + survey_month + survey_year + station_id + nearest_dist~ paste0("month_", offset),
      value.var = "heat_flag",
      fun.aggregate = max
    ) %>%
  setcolorder(c("hospital_id", "survey_month", "survey_year", 'station_id', 'nearest_dist', paste0("month_", 0:24))) 

write.csv(heat_summary_matrix_arb, r"(./data/heat_matrix_dayCriteria.csv)")
```

Display a snippet of the arbitrary heat summary matrix.

```{r arb-summary-matrix}
kable(head(heat_summary_matrix_arb),'html') %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(height = "300px")
```


## Map {#heat-map}

The South Okanagan General Hospital in British Columbia experienced a heat event, as defined by Environment and Climate Change Canada, in the 18th month prior to the survey administration month. The Hospital is located in the Southeastern British Columbia zone, where the daytime max temperature threshold is 35C and the nighttime minimum threshold temperature is 18C. The weather station and hospital locations are presented below. 

```{r heat-map, message=FALSE}
#get hospital 509, heat event in month 18
hospital_509 <- filter(survey_sf_eccc, hospital_id == 509) %>% 
  mutate(heat_event = "Month 18") %>% st_transform(crs=st_crs(4326))

#get station 1039 since it is the close one
station_of_interest <- weather_dl(station_id = c(1039), interval='day', start='2018-01-01', end='2020-01-31') %>% 
  slice(1) %>%
  st_as_sf(coords = c("lon","lat"), crs=st_crs(4326)) %>% st_transform(crs=st_crs(4326))

#create the map elements
mapview(
  filter(hospital_509),
  col.regions = "red",
  cex = 6,
  layer.name = "Hospital Location",
  popup = leafpop::popupTable(
    hospital_509,
    zcol = c("odhf_facility_type", "hospital_id", "postalcode", "survey_year", "survey_month", "heat_event")
  )
) +
mapview(
  station_of_interest,
  col.regions = "blue",
  alpha.regions = 1,
  layer.name = "Closest Station",
  popup = leafpop::popupTable(
    station_of_interest,
    zcol = c("station_name", "station_id")
    )
)
```

