```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/matth/Climate_and_Health/bookdown_project')
```

# Environmental Exposures {#environmental-exposures}

## Introduction {#introduction-environmental-data}

Sample data is Canadian public hospital data. Environmental events can have an effect disproportionate to their size on a variety of different points of interest. For this reason, a 25km buffer is established around each point for all the environmental exposures except for heat.  

```{r load-hospital-data}
odhf_data <- read.csv("./data/odhf_hospitals.csv", row.names = 1)
```

## Wildfires {#wildfires}

### Environmental Data Preparation {#fire-data-preparation}

Data can be downloaded for free from the [Canadian Wildland Fire Information System](#../index.Rmd/datasets "Datasets"). To begin with the wildfire data, we read the NBAC wildfire polygons (1972-2024) from the .shp file available from CWFIS.

-   Tip: Shapefiles are a set of files. Ensure the .shp, .shx, .dbf, and (ideally) .prj sit together in the same folder.

```{r fire-load-and-limit-data, eval=FALSE}
#TODO create unified language and narrative view in html text throughout
#read in the NBAC data
poly_fires <- st_read(r"(./data/FireData/NBAC_1972to2024_20250506.shp)") 

#look at the data
str(poly_fires)

#normalise column names
names(poly_fires) <- tolower(names(poly_fires)) 

#Given the extremely large size, it's best to reduce the dataset to include only the years required
poly_fires <- poly_fires %>% 
  filter(year >= 2015)

#limit to a selection of columns that are relevant
poly_fires <- select(poly_fires, 
                     nfireid, basrc,
                     year, 
                     ag_sdate, #agency start date as date format
                     ag_edate, #agency end date as date format
                     poly_ha, 
                     admin_area,
                     geometry)

#rename columns for easier understanding
poly_fires <- poly_fires %>%
  mutate(
    start_day = day(ag_sdate),
    start_month = month(ag_sdate),
    start_year = year(ag_sdate),
    end_day = day(ag_edate),
    end_month = month(ag_edate),
    end_year = year(ag_edate),
    start_date = ag_sdate, 
    end_date = ag_edate,
    size = poly_ha) %>%
  select(-poly_ha)  #remove the original poly_ha column

#write the newly formed dataset to memory, it will be loaded and used in this example
st_write_parquet(poly_fires, r"(./data/wildfire_data_sf.parquet)")

names(poly_fires)
```
```{r fire-data-pickup, include = FALSE}
#read the fire data snippet for reduced memory constraints
poly_fires <- st_read_parquet(r"(./data/wildfire_data_sf.parquet)")
```

### Compute Intersections {#fire-intersections}

We need to compute the intersections between the fires and buffered hospitals. In this example, we select only fires greater than 200 hectares for computing intersections using EPSG 3978. 

```{r fire-odhf-intersection}
min_fire_size <- 200 #in ha
buffer_distance = 25000 #in metres
poly_fires <- poly_fires %>% filter(size > min_fire_size) #filter fires by size

#use hospital data for fires
odhf_fires <- odhf_data %>% 
  mutate(long = as.numeric(long)) %>% 
  filter(!is.na(lat) & !is.na(long)) #filter out na points

odhf_fires_sf <- odhf_fires %>% 
  st_as_sf(coords=c("long", "lat"), crs = st_crs(4326))

odhf_fires_sf <- st_transform(odhf_fires_sf, st_crs(poly_fires)) #transform to a metres pcs

#Create buffer around each location
odhf_buffers <- st_buffer(odhf_fires_sf, dist = buffer_distance) #create the buffer in metres

#intersection matrix, the product refers to row indices, not id numbers
fire_matches <- st_intersects(odhf_buffers, poly_fires) #run st_intersection to get geometries, but takes way longer
```

Link the intersection matrix back to the original dataframe.

```{r fire-link-intersections}
#create the intersection dataframe from the intersection matrix
intersection_df <- map2_dfr(seq_along(fire_matches),  fire_matches,
  ~ tibble(
    hospital_id = odhf_fires_sf$hospital_id[.x], #.x indicates it's from the sequence (argument 1)
    n_fire_id = poly_fires$nfireid[.y], #.y indicates it's from poly_fires (argument 2)
    survey_year = odhf_fires_sf$survey_year[.x],
    survey_month = odhf_fires_sf$survey_month[.x],
    fire_start_date = poly_fires$ag_sdate[.y],
    fire_end_date = poly_fires$ag_edate[.y],
    intersects = 1
  )
)

#create useful date fields for the 24-month date lag
intersection_df$survey_eom_date <- ymd(paste(intersection_df$survey_year, intersection_df$survey_month, "01", sep = "-")) + months(1) - days(1)
intersection_df$survey_25_prev <- ymd(paste(intersection_df$survey_year, intersection_df$survey_month, "01", sep = "-")) -months(24)
intersection_df$fire_start_date <- as.Date(intersection_df$fire_start_date, format="%YYYY-%MM-%DD")#make into a date
intersection_df$fire_end_date <- as.Date(intersection_df$fire_end_date, format="%YYYY-%MM-%DD")#make into a date
```

We also need to deal with extraneous cases like rows with no fire end dates and those that were recorded as instantaneous (same start and end date).

```{r fire-date-range-coverage}
#this gets all overlapping dates, treating rows with no fire end date as one day instantaneous events
intersection_overlap <- intersection_df %>%
  mutate(
    #define effective fire range: point if no end date, range otherwise
    fire_start_eff = fire_start_date,
    fire_end_eff   = coalesce(fire_end_date, fire_start_date)
  ) %>%
  filter(
    #survey range ovelraps with fire range
    survey_25_prev <= fire_end_eff,
    survey_eom_date >= fire_start_eff
  )
```

### Date Lag {#fire-date-lag}

We need to determine the 24 months prior to each survey date at each hospital. Hospitals were surveyed at different times, so the date of the survey will be recorded along with each month from the 0th to 24th month prior (survey month, survey month - 1, survey month - 2, etc.).

```{r fire-date-lag}
#expand each survey row into last 24 months
expand_months <- function(d, hospital_id, survey_month, survey_year) {
  data.table(
    hospital_id = hospital_id,
    survey_year = survey_year,
    survey_month = survey_month,
    offset = 0:24,
    date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
  )
}

#apply the function
#identifies how many months a given month is offset from the 0th month for that hospitals survey date
odhf_fires_sf$dates <- ymd(paste(odhf_fires_sf$survey_year, odhf_fires_sf$survey_month, "01", sep = "-")) + months(1) - days(1)
#for unique examples (one hospital-survey combo)
fires_un <- unique(odhf_fires_sf, by = c("hospial_id","dates"))
fires_un <- as.data.table(fires_un)
survey_expanded <- fires_un[, expand_months(dates, hospital_id, survey_month, survey_year), by = .(hospital_id, dates)] #expand the months backwards for each unique hospital_id/date
survey_expanded <- survey_expanded %>% select(-3) #get rid of the duplicate column
```

### Binary Matrix {#fire-summary-matrix}

Create a binary matrix for each hospital-survey date combination for each of the 24 months.

```{r fire-summary-matrix}
#create a detailed data frame of all intersections and non-intersections, regardless of date, as an intermediary product 
fires_detailed <- survey_expanded %>%
  left_join(intersection_df, by = c("hospital_id", "survey_year", "survey_month")) %>% #join the data on year month and id
  mutate(
    month_ym = ymd(paste0(date, "01")), #set the date
    month_start = month_ym, #get the first month as a date
    month_end = ceiling_date(month_ym, "month") - days(1), #get the last month as a date
    fire_start_eff = fire_start_date, #all have start date or nothing
    fire_end_eff = coalesce(fire_end_date, fire_start_date), #coalesce if nothing 
    intersect = case_when(
      is.na(fire_start_date) ~ 0,  #fire had no date or there is no fire
      month_start <= fire_end_eff & month_end >= fire_start_eff ~ 1, 
      TRUE ~ 0
    )
  ) %>%
  select(hospital_id, survey_year, survey_month, offset, date, #select these columns
         intersect, n_fire_id, fire_start_date, fire_end_date, fire_start_eff, fire_end_eff)

#create binary matrix
fires_summary_matrix <- fires_detailed %>% 
  #filter(intersect == 1) %>% #if only interested in the results that have intersections
  group_by(hospital_id, survey_year, survey_month, offset) %>%  #group by hospital id
  right_join(survey_expanded, by=c("hospital_id", "survey_year", "survey_month", "offset")) %>% #join
  mutate(intersect = replace_na(intersect, 0)) %>% #make sure there are no NAs
  ungroup() %>%
  as.data.table() %>% #convert to data table for some easier mutability
  dcast(., #reshape data to wide format based on preceding months
      hospital_id + survey_month + survey_year ~ paste0("month_", offset),
      value.var = "intersect",
      fun.aggregate = max
    ) %>%
  setcolorder(c("hospital_id", "survey_month", "survey_year", paste0("month_", 0:24))) #column order makes sense

#create an output sample to display
fires_summary_matrix[c(120:130), ] %>% 
  kbl(format = "html", caption = "Fire Summary Matrix") %>%
  kable_styling() %>%
  scroll_box(width = "100%")

#write csv result
write.csv(fires_summary_matrix, r"(./data/fires_presence_absence.csv)")

```

## Floods {#floods}

### Environmental Data Preparation {#floods-data-preparation}

Natural Resources Canada provides the flood archive data set through [Open Canada](#../index.Rmd/datasets "Datasets"). The data is provided in bulk as a .gdb file with several layers. We are specifically interested in the 'EGS_Flood_Product_Archive' layer.

```{r floods-initial-data, warning=FALSE}
#load raw flood data
gdb_path <- "./data/EGS_Flood_Product_Archive.gdb"
st_layers(gdb_path)
flood_data <- st_read(gdb_path, layer = "EGS_Flood_Product_Archive", options = "METHOD=SKIP")
```

### Data Cleaning {#floods-data-cleaning}

We are interested in floods that happened within 25km of the POI survey data. After converting the survey data to the same coordinate reference system as the flood data we can look at the intersection of the two data points.

```{r floods-survey-data-preprocess}
#need to make some data to verify
#names throw an error due to character incompatibility, so remove altogether
odhf_valid <- odhf_data
survey_sf <- odhf_valid %>% st_as_sf(coords=c("long", "lat"), crs=st_crs(4326)) %>% select(-facility_name) %>% st_transform(st_crs(flood_data))
```

### Data Manipulation {#floods-data-manipulation}

Buffer distance is easily customisable for different use cases. This chunk will loop through each of the points of interest in the hospital dataset to determine flood exposure within the two years prior to each hospitals survey.

```{r floods-main-loop}
buffer_distance = 25000 #in metres

#create a buffer around each poi
survey_buffered <- st_buffer(survey_sf, dist = buffer_distance)

#find floods within each poi's buffer
flood_matches <- st_intersects(survey_buffered, flood_data)

#extract year and month from flood event datetime
flood_data <- flood_data %>%
  mutate(flood_year = year(datetime_utc),
         flood_month = month(datetime_utc))

#initialize an empty data frame to store results
exposure_matrix <- matrix(0, nrow = nrow(survey_sf), ncol = 25)
colnames(exposure_matrix) <- paste0("month_", 0:24)  #column names for each month

#loop through each poi and determine flood exposure
for (i in seq_len(nrow(survey_sf))) {
  #extract the survey date from the survey related columns
  survey_year <- survey_sf$survey_year[i]
  survey_month <- survey_sf$survey_month[i]
  
  #compute the 24-month range
  month_seq <- seq.Date(from = as.Date(sprintf("%d-%02d-01", survey_year, survey_month)) %m-% months(1), 
                        length.out = 24, by = "-1 month")
  month_seq_df <- data.frame(year = year(month_seq), month = month(month_seq))
  
  #get the indices of intersecting floods for this poi
  flood_indices <- flood_matches[[i]]
  
  if (length(flood_indices) > 0) {  #if floods exist within the buffer
    flood_events <- flood_data[flood_indices, c("flood_year", "flood_month")]
    
    #check for each of the past 24 months if a flood occurred
    for (j in 1:24) {
      target_year <- month_seq_df$year[j]
      target_month <- month_seq_df$month[j]
      
      #if a flood occurred in this year-month, set exposure to 1
      if (any(flood_events$flood_year == target_year & flood_events$flood_month == target_month)) {
        exposure_matrix[i, j] <- 1
      }
    }
  }
}

#convert matrix to a data frame and merge with survey data
exposure_df <- as.data.frame(exposure_matrix)

#bind to the presence absence matrix
survey_floods_pa <- cbind(survey_sf, exposure_df)

#summarise based on the number of unique hospitals
survey_floods_pa %>% summarise(n_unique = n_distinct(hospital_id)) 

#create a summary dataset of hospitals with only flood events for easier viewing
survey_floods_pa_filtered <- survey_floods_pa %>%
  filter(if_any(starts_with("month_"), ~ . == 1))

#print a selection of data
kable(head(survey_floods_pa_filtered),'html') %>%
  kable_styling(full_width = FALSE) %>%
  scroll_box(height = "300px")

#write to a directory 
write.csv(survey_floods_pa, r"(./data/flood_presence_absence.csv)")
```

## Hurricanes {#hurricanes}

### Environmental Data Preparation {#hurricanes-data-preparation}

Hurricane data was sourced from the National Oceanic and Atmospheric Administration (NOAA). Data can be obtained programmatically (as below), or manually for each hurricane of interest from the [NOAA website](https://www.nhc.noaa.gov/gis/archive_besttrack.php?year= "Hurricane Best Tracks"). Should you choose to download the files yourself, the script is designed to use shapefiles (.shp).

-   It is important to note that the following script should be used sparingly

We first need to specify when and where the hurricane data is retrieved from.

```{r hurricanes-set-data-extraction-range}
base_page <- "https://www.nhc.noaa.gov/gis/archive_besttrack.php?year="
base_zip  <- "https://www.nhc.noaa.gov/gis/best_track/"
hurricane_years_interest <- 2008:2025 #getting data for all years from 2008 to 2025
```

Next, we set up the storage location for the hurricane best track files and create a lookup table for later. It is important to set a viable output directory for the best tracks from the website.

```{r hurricanes-define-output}
#set an output directory for the data to download into
base_shape_dir <- "./data/HurricaneRawData/nhc_best_tracks"
dir.create(base_shape_dir, showWarnings = FALSE)

#create a lookup table for the storms that can be used later to create a connection between the storm name and data
storm_summary <- data.frame(Year = integer(), Storms = integer())
storm_lookup  <- data.frame(year = integer(), storm_id = character(), storm_name = character())
```

Finally, we get the files from the website for each year specified. Within this snippet, a lookup table for the file names and actual hurricane name is created and saved as a .csv for later reference. (TODO: add a switch to get just the names list vs. the file and names list)

```{r hurricanes-scrape-data, eval = FALSE}
#process every year for the year range specified above
for (year in hurricane_years_interest) {
  cat("processing", year, "\n")
  
  #read the page, catch if the response is blank
  page <- tryCatch({
    read_html(paste0(base_page, year))
  }, error = function(e) {
    cat("could not read page for", year, "\n")
    return(NULL)
  })
  if (is.null(page)) next
  
  #extract all the rows on the page to get the hurricane codes and names from the table for that year
  rows <- html_nodes(page, "tr")
  
  #loop through rows to match "al##" and extract name from second column to create lookup table
  for (row in rows) {
    cols <- html_nodes(row, "td") #get all the td html elements
    
    if (length(cols) >= 2) {
      id_text <- str_trim(html_text(cols[1])) #just want the id from the td elements
      
      if (str_detect(tolower(id_text), "^al\\d{2}$")) { #regex to get the atlantic (al) storm and the id number
        name_text <- str_trim(html_text(cols[2])) #get the nominal identifier
        storm_lookup <- rbind( #collect and add to the storm loookup table
          storm_lookup,
          data.frame( #create a data frame with the year, storm id from the td element, and the storm name
            year = year,
            storm_id = tolower(id_text),
            storm_name = name_text,
            stringsAsFactors = FALSE
          )
        )
      }
    }
  }
  
  #get just the text from all the td html elements
  td_text <- html_text(html_nodes(page, "td"), trim = TRUE)
  
  #get the strom ids again (for redundancies sake)
  storm_ids <- unique(str_to_lower(str_subset(td_text, "^al\\d{2}")))
  
  #make sure it's just the 4 character id with more regex
  storm_ids <- gsub("\\s.*", "", storm_ids)
  
  cat(" found", length(storm_ids), "atlantic storms\n")
  
  #storm summary for creating storm ids in file names
  storm_summary <- rbind(storm_summary,
                         data.frame(Year = year, Storms = length(storm_ids)))
  
  #file naming system is regular within the website hierarchy, so download all zip shapefiles using the engineered storm names
  for (storm_id in storm_ids) {
    zip_name <- paste0(storm_id, year, "_best_track.zip")
    url <- paste0(base_zip, zip_name)
    dest <- file.path(base_shape_dir, zip_name) #set outpath based on string specified earlier
    
    cat(" downloading:", zip_name, "\n")
    resp <- tryCatch({
      GET(url, write_disk(dest, overwrite = TRUE)) #get zip file from constructed url
    }, error = function(e) {
      cat(" failed to download", zip_name, "\n") #sometimes there are no files to download, especially in earlier years
      return(NULL)
    })
    
    #unzip the files too, easier to do now than later
    if (!is.null(resp) && status_code(resp) == 200) {
      unzip(dest, exdir = file.path(base_shape_dir, paste0(storm_id, "_", year)))
    } else {
      cat(" file not found:", url, "\n")
    }
    #be kind to the server and don't get banned from scraping
    Sys.sleep(1.1)
  }
}

write.csv(storm_lookup, r"(./data/HurricaneRawData/storm_lookup_table.csv)")
```

### Data Cleaning {#hurricanes-data-cleaning}

Now that the data is downloaded and ready, it is time clean it up. There are two files of interest for the hurricanes, the point files and the polygon (radii) files. These files need to be manipulated so that dates and ids are consistent between the two file sets and between years.

```{r hurricanes-map-months}
#standardizing naming conventions in files
month_map <- c(
  "jan" = 1, "january" = 1, "01" = 1,
  "feb" = 2, "february" = 2, "02" = 2,
  "mar" = 3, "march" = 3, "03" = 3,
  "apr" = 4, "april" = 4, "04" = 4,
  "may" = 5, "05" = 5,
  "jun" = 6, "june" = 6, "06" = 6,
  "jul" = 7, "july" = 7, "07" = 7,
  "aug" = 8, "august" = 8, "08" = 8,
  "sep" = 9, "september" = 9, "09" = 9,
  "oct" = 10, "october" = 10, "10" = 10,
  "nov" = 11, "november" = 11, "11" = 11,
  "dec" = 12, "december" = 12, "12" = 12
)

#base directory where files are saved
base_dir <- r"(./data/HurricaneRawData/nhc_best_tracks)"

#set an ID for all of the files in the pts files and radii files
#get all shapefiles recursively
shp_files <- list.files(base_dir, pattern = "(_pts\\.shp$|_radii\\.shp$)", recursive = TRUE, full.names = TRUE)
```

Add the storm ID for each storm to the file names and overwrite the shapefile.

```{r hurricanes-add-storm-id, warning=FALSE, message=FALSE, results='hide', eval=FALSE}

#process the files
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  #read the shapefile
  shp <- st_read(shp_path, quiet = TRUE)
  
  #extract part before first underscore
  file_prefix <- sub("_.*", "", basename(shp_path))
  
  #add the storm id file
  shp <- shp %>% mutate(storm_id = file_prefix)
  
  #overwrite shapefile
  st_write(shp, shp_path, delete_dsn = TRUE, quiet = TRUE)
}

```

Date and time need to be consistent between the point and polygon files.

```{r hurricanes-alter-points-files, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
#Create a new directory for processed files
output_dir <- file.path(base_dir, "processed")

#processing for just the _pts.shp files to normalize dates and times
base_dir <- r"(./data/HurricaneRawData/nhc_best_tracks)"
output_dir <- file.path(base_dir, "processed")
dir.create(output_dir, showWarnings = FALSE)

#get all the shapefiles recursively
shp_files <- list.files(base_dir, pattern = "_pts\\.shp$", recursive = TRUE, full.names = TRUE)

#recursively alter the pts files to the normalized date and time fields with ids
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  shp <- st_read(shp_path, quiet = TRUE)
  
  #files had different formats for dates and times, this normalizes them
  #important step to combine everything for later analysis
  shp <- shp %>%
    mutate(
      mm = str_pad(as.integer(unname(month_map[tolower(MONTH)])), 2, pad = "0"),
      dd = str_pad(as.integer(DAY), 2, pad = "0"),
      hh = str_sub(HHMM, 1, 2),
      event_id = tolower(paste0(storm_id, mm, dd, hh))
    )
  
  #preserve folder structure in output
  rel_path <- path_rel(shp_path, start = base_dir)
  rel_path <- sub("^processed/", "", rel_path)
  out_path <- file.path(output_dir, rel_path)
  dir_create(dirname(out_path))
  
  #write output
  st_write(shp, out_path, delete_dsn = TRUE, quiet = TRUE)
}
```

The points files also need to be done in a similar way.

```{r hurricanes-alter-radii-files, message=FALSE, warning=FALSE, results='hide', eval=FALSE}
#slightly different process for the _radii file they already have the synoptime field and those just need to be modified
shp_files <- list.files(base_dir, pattern = "_radii\\.shp$", recursive = TRUE, full.names = TRUE)

#recursively alter the pts files to fie normalized date and time fields with ids
for (shp_path in shp_files) {
  message("Processing: ", shp_path)
  
  shp <- st_read(shp_path, quiet = TRUE)
  
  #create matching event_id to the points fields so they can be related (since they have different kinds of information)
  shp <- shp %>%
    mutate(
      event_id = tolower(
        paste0(storm_id, substr(SYNOPTIME, 5, nchar(SYNOPTIME))) #this will be storm id, date, and UTC hour in 24 hour format i.e. [al01][20250831][12]
      )
    )
  
  #different kind of id, radii refers to the range of wind at that knot level i.e. _34 would be 34 knot wind
  shp <- shp %>%
    mutate(
      event_r_id = tolower(
        paste0(storm_id, substr(SYNOPTIME, 5, nchar(SYNOPTIME)), "_", RADII)#this will be storm id, date, and UTC hour in 24 hour format, and radii distance  i.e. [al01][20250831][12][_34]
      )
    )
  
  #preserve folder structure
  rel_path <- path_rel(shp_path, start = base_dir)
  out_path <- file.path(output_dir, rel_path)
  dir_create(dirname(out_path))
  
  #write output
  st_write(shp, out_path, delete_dsn = TRUE, quiet = TRUE)
}
```

### Data Manipulation {#hurricanes-data-manipulation}

Now that the hurricane files are prepared properly, the associated points data of interest, (in this case hospitals), can be used within the files. The point files in this example have a series of months and years in columns when surveys were conducted, a latitude and longitude column where we can identify

First we load all the storms.

```{r hurricanes-load-storms, warning=FALSE}
#where all the processed files are
storms_dir <- r"(./data/HurricaneRawData/nhc_best_tracks/processed)"

#get all the radii shapefiles
#I settled on radii, because its the reach of the wind
cols <- c("event_id", "event_r_id", "storm_id")

storms_list <- purrr::map(shp_files, function(x) {
  st_read(x, quiet = TRUE) %>%
    select(any_of(cols))
})

#get all the storms
all_storms <- dplyr::bind_rows(storms_list)
all_storms <- st_transform(all_storms, 4326)
all_storms <- st_make_valid(all_storms)

all_storms$months <- ym(str_sub(all_storms$event_id, 5, 10))
```

Now that all the data is loaded and ready, the intersection with all the points of interest needs to be performed and a number of data outputs are created.

1.  A long table of storms and intersections with hospitals.
2.  A wide table with the ID's of the surveyed locations and a matrix of 1's and 0's identifying if they were impacted by a storm during the previous 24 months.
3.  A table with daily intersection information for the month the survey was conducted.
4.  A long table with only intersection information

```{r hurricanes-main-loop}
#the website of interest for tracks
website <- 'https://www.nhc.noaa.gov/gis/archive_besttrack.php'

#lookup table with names that was created in an earlier step
storm_names_dir <- r"(./data/HurricaneRawData/storm_lookup_table.csv)"
##############################################################
#postcode files should have matching column titles to do this in a loop:
#id column
#long, lat
#survey_year, survey_month
#make sure that these column names exist
long_col <- "long"
lat_col <- "lat"
survey_id_col <- "hospital_id"
survey_year_col <- "survey_year"
survey_month_col <- "survey_month"
#############################################################
cols <- c("geometry", "event_id", "event_r_id", "storm_id")
#buffer distances in metres to go around point of interest
bufs <- c(25000)#, 0, 50000, 100000)
#years of survey dates for table naming conventions
years <- c('all')
#this is paried with the above years, should match up!
survey_hospitals <- r"(./data/odhf_hospitals.csv)"
dirs <- c(survey_hospitals)
#results storage
hurricane_results <- list()


#this is a double loop, the first one goes through each survey year of interest 
for (k in seq_along(years)){
  message("Naming suffix: ", years[k])
  #the second loop applies the buffer size to the survey data for intersection analysis 
  #between storm pathway and survey data
  for (i in seq_along(bufs)){
    message("Point buffer extent: ", bufs[i]/1000, " km")
    #read the survey data
    survey_data <- read.csv(dirs[k])
    #convert the survey data to an sf object for the intersection
    survey_sf <- st_as_sf(
      #eliminate the Nas
      survey_data[!is.na(survey_data[[long_col]]) & !is.na(survey_data[[lat_col]]), ],
      coords = c(long_col, lat_col),
      crs = 4326
    )
    #make sure its the appropriate crs
    survey_sf <- st_transform(survey_sf, 4326)
    #apply the buffer to the survey points
    survey_buffer <- st_buffer(survey_sf, bufs[i])
    #create a date type column from the survey year and survey month columns
    survey_buffer$months <- ym(paste0(survey_buffer[[survey_year_col]], survey_buffer[[survey_month_col]]))
    survey_buffer$survey_id <- paste(survey_buffer[[survey_id_col]], survey_buffer[[survey_month_col]], survey_buffer[[survey_year_col]], sep="_")
    #get all storms that happen within 30 months of the earliest survey date and the latest survey month of that year
    #this could be a pretty variable range of dates
    all_storms_within_range <- all_storms %>% filter(months >= (min(survey_buffer$months)-months(30)) & months <= max(survey_buffer$months))
    #do the intersection between the surveys and all of those storms
    intersection <- st_intersects(survey_buffer, all_storms_within_range, sparse = TRUE)
    
    #build the dataframe
    intersection_df <- map2_dfr(
      #for each intersection
      seq_along(intersection),
      #create a little tibble of point, strom, survey, and if it intersects
      #creates an association between each poi and storm intersect event and attaches the date of the survey
      intersection,
      ~ tibble(
        point_id = survey_sf[[survey_id_col]][.x],
        storm_id = all_storms_within_range$event_r_id[.y],
        survey_date = survey_buffer$months[.x],
        intersects = 1
      )
    )
    
    #get the strom names
    storm_lookup <- read.csv(storm_names_dir, sep=",")
    
    #preserve the intersection dataframe
    matrix_df <- intersection_df
    #get the survey data anew
    survey_data <- read.csv(dirs[k])
    #set blanks to na's
    survey_data[survey_data == ""] <- NA
    #only keep non na points
    survey_data <- survey_data[!is.na(survey_data[[long_col]]) & !is.na(survey_data[[lat_col]]), ]
    #convert to data.table
    survey_data <- as.data.table(survey_data)
    #organize dates
    survey_data[, dates := ceiling_date(make_date(survey_year, survey_month, 1), "month") - days(1)]
    
    #convert to data.table
    matrix_df <- as.data.table(matrix_df)
    #only interested in the intersections, there shouldn't be any misses, but just in case
    matrix_df <- matrix_df[intersects == 1]
    #set the date column in the matrix_df data frame
    matrix_df[, year_month := format(ym(str_sub(storm_id, 5, 10)), "%Y%m")]
    #only need to know if there is an intersection on that month, 
    #so get unique entries for each poi and month combination
    matrix_df <- unique(matrix_df, by = c("point_id","survey_date", "year_month"))
    
    #expand each survey row into last 24 months (month_0 = survey month, month_1 = survey-1, etc)
    expand_months <- function(d, s_id, survey_month, survey_year) {
      data.table(
        survey_id = s_id,
        survey_year = survey_year,
        survey_month = survey_month,
        offset = 0:24,
        date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
      )
    }
    
    #apply the function
    #identifies how many months a given month is offset from the 0th month for that pois survey date
    survey_expanded <- survey_data[, expand_months(dates, get(survey_id_col), survey_month, survey_year), by = c(survey_id_col, survey_year_col, survey_month_col)]
    survey_expanded$survey_date <- ym(paste0(survey_expanded[[survey_year_col]], survey_expanded[[survey_month_col]]))

    #join the storms on survey id and the date at year_month
    result <- matrix_df[survey_expanded, 
                        on = c("point_id" = survey_id_col, "year_month" = "date")]
    
    #add 0's to the matrix of 1's and 0's
    result[is.na(intersects), intersects := 0]
    
    #join storm names for easier identification
    storm_lookup <- as.data.table(storm_lookup)
    #storm id and year to cross reference with website.
    storm_lookup[, storm_id_year := paste0(storm_id, year)]
    result[, storm_id_year := str_sub(storm_id, 1, 8)]
    result <- merge(
      result,
      storm_lookup[, .(storm_id_year, storm_name)],
      by = "storm_id_year",
      all.x = TRUE
    )
    
    #pivot to one row per id and columns month_0 to 24\
    formula_str <- paste("survey_id + survey_month + survey_year ~ paste0('month_', offset)")
    result_months <- dcast(
      result,
      as.formula(formula_str),
      value.var = "intersects",
      fun.aggregate = max
    )
    
    #join the lat and long from the original dataset back for joining later
    on_vec <- setNames(c(survey_id_col, "survey_year", "survey_month"), 
                   c("survey_id", "survey_year", "survey_month"))

    result_months[survey_data, 
                  on = on_vec,
                  `:=`(lat = i.lat, long = i.long)]
    
    ###################month 0 information day by day ###############################
    matrix_df <- intersection_df
    matrix_df <- matrix_df %>% 
      #only intersections
      filter(intersects == 1) %>%
      #get date to the day of the storm
      mutate(year_month_day = ymd(str_sub(storm_id, 5, 12))) %>% 
      mutate(storm_id_long = str_sub(storm_id, 0, 12)) %>% 
      #one row per day, doesn't matter the wind speed
      distinct(., point_id, year_month_day,.keep_all=TRUE) %>% 
      #format date as a date again
      mutate(year_month_day = format(year_month_day, "%Y-%m-%d"))
    
    #get survey data  
    survey_data <- read.csv(dirs[k])
    #set blanks to na's
    survey_data[survey_data == ""] <- NA
    #only keep non na points
    survey_data <- survey_data[!is.na(survey_data[[long_col]]) & !is.na(survey_data[[lat_col]]), ]
    #assume survey was conducted on the final day of the month, so check every day of the month
    survey_data$dates <- ceiling_date(make_date(year=survey_data[[survey_year_col]], month = survey_data[[survey_month_col]], day = 1), "month") - days(1)
    
    #expand to every day of the month
    expand_survey_data <- survey_data %>%
      rowwise() %>%
      mutate(days_in_month = list(seq(from = floor_date(dates, "month"), to = dates, by = "1 day"))) %>%
      unnest(days_in_month) %>%
      rename(date = days_in_month) %>% 
      mutate(date = format(date,"%Y-%m-%d"))
    
    #join the two datasets of storm data and survey data
    result_days <- left_join(expand_survey_data, matrix_df, by = c("date" = "year_month_day", setNames("point_id", survey_id_col)))
    
    #add to final dataframe of dataframes
    key <- paste0("result_", bufs[i]/1000, "k_", years[k])
    hurricane_results[[key]] <- list(
      long = result,
      wide = result_months,
      daily = result_days,
      matrix = matrix_df
    )
  }
}

#get the hurricane matrix wide data
hurricane_matrix <- hurricane_results[["result_25k_all"]][["wide"]] %>%
  relocate(all_of(paste0("month_", 0:24)), .after = last_col())

#display
hurricane_matrix[c(278:290), ] %>% 
  kbl(format = "html", caption = "Hurricane Summary Matrix") %>%
  kable_styling() %>%
  scroll_box(width = "100%")

#write the matrix
write.csv(hurricane_matrix, r"(./data/hurricane_presence_absence.csv)")

```

## Heat {#heat}

### Environmental Data Preparation {#heat-data-preparation}

Heat data was sourced from the `weathercan` R package.

```{r}
odhf_valid <- odhf_data
survey_sf <- odhf_valid %>% st_as_sf(coords=c("long", "lat"), crs=st_crs(4326)) %>% select(-facility_name)
survey_sf <- st_transform(survey_sf, 3978)
stations_list <- stations() 
names(stations_list)
```

In order to collect the relevant weather stations and time frames, the minimum and maximum survey data collection periods needed to be collected from the ODHF data.

```{r}
survey_sf <- survey_sf %>%
  mutate(
    weather_start = make_date(survey_year - 2, survey_month, 1),  #Start date: 2 years before
    weather_end = make_date(survey_year, survey_month, days_in_month(make_date(survey_year, survey_month, 1))))

min_start_year <- min(year(survey_sf$weather_start), na.rm = TRUE)
max_end_year <- max(year(survey_sf$weather_end), na.rm = TRUE)

stations_by_day <- filter(stations_list, 
                                      interval %in% c("hour","day"), 
                                      #setting the min coverage years for data at the stations
                                      #is to increase likelihood that the same hospital location will be assigned                                          to the same weather station
                                      #across survey years 
                                      #this will reduce noise introduced as a result of pulling data for the same                                         location from different sources 
                                      #which may have different data available, different frequencies of                                                 collection, etc. 
                                      start <= min_start_year, 
                                      end >= max_end_year)
```

Next, we need to get the nearest distance to stations for evaluation of if they are too close or too far away. In the case of this example, no station will be too far away.

```{r}
stations_by_day_sf <- st_as_sf(stations_by_day, coords=c("lon", "lat"), crs = 4326)
stations_by_day_sf <- st_transform(stations_by_day_sf, st_crs(survey_sf))
nearest_stations <- st_nearest_feature(survey_sf, stations_by_day_sf)
survey_sf$nearest_dist <- st_distance(survey_sf, stations_by_day_sf[nearest_stations,], by_element = TRUE)
survey_sf$near_station_id <- stations_by_day_sf[nearest_stations,]$station_id
survey_sf$station_interval <- stations_by_day_sf[nearest_stations,]$interval
survey_sf <- survey_sf %>% rename(station_id=near_station_id)
```

Get the data.

```{r get-weather-data, eval=FALSE}
hourly_station_ids <- filter(survey_sf, station_interval == "hour") %>% pull(station_id)
daily_station_ids <- filter(survey_sf, station_interval == "day") %>% pull(station_id)
daily_weather <- weather_dl(station_ids = daily_station_ids, start=min(survey_sf$weather_start), end=max(survey_sf$weather_end), interval='day')
hourly_weather <- weather_dl(station_ids = hourly_station_ids, start=min(survey_sf$weather_start), end=max(survey_sf$weather_end), interval='hour')
```

Group and combine.

```{r group-weather-data, eval=FALSE}
hourly_as_daily_weather_original <- hourly_weather %>% 
  group_by(station_id, date, lat, lon) %>% 
  summarise(
    min_temp = min(temp, na.rm = FALSE),
    max_temp = max(temp, na.rm = FALSE),
    .groups = "drop"
  )
hourly_as_daily_weather <- subset(hourly_as_daily_weather_original, select = -c(lat, lon))
combined_hourly_daily <- bind_rows(hourly_as_daily_weather, daily_weather) %>%
  select(any_of(intersect(names(hourly_as_daily_weather), names(daily_weather))))
write_parquet(combined_hourly_daily, r"(./data/odhf_weather_stations.parquet)")
```

### Heat ECCC {#heat-daily-eccc}

```{r heat-combined-daily-eccc}
combined_hourly_daily <- read_parquet(r"(./data/odhf_weather_stations.parquet)")
eccc_friendly <- st_read(r"(./data/roughEcccMap_final.gpkg)")
survey_sf_eccc <- st_join(survey_sf, eccc_friendly)

consecutive_days <- 2 #2 days and nights as the criteria.

#test <- left_join(hospital_sf_eccc, combined_hourly_daily, by='station_id')

heat_flagged_eccc <- survey_sf_eccc %>%
  left_join(combined_hourly_daily, by='station_id') %>%  #date, station_id, max_temp, min_temp 
  unique(.,by=c("hospital_id", "survey_month", "survey_year", "date")) %>%  #get unique entries based on survey_id, survey month + year, and date of station data collection
  arrange(station_id, date) %>% #need dates in order chronologically
  group_by(station_id, hospital_id, survey_month, survey_year) %>% #station_ids grouped together
  mutate(
    hot = max_temp >= day_thresh & min_temp >= nite_thres, #if the max_temp is greater than the day threshold AND the minimum temperature is greater than the nighttime threshold
    run_id = rleid(hot) #add unique id for runs
  ) %>% 
  group_by(hospital_id, survey_month, survey_year, station_id, run_id) %>% #group by station id and the run ids (looking for two or more)
  mutate(
    run_len = sum(hot), #sum the number of hot events per run                                    
    heat_event = hot & run_len >= consecutive_days #only classed as a heat event if two or more days       
  ) %>%
  ungroup() %>%
  select(-run_id, -run_len) %>% 
  mutate(date = as.Date(date))
```

#### Date Lag {#heat-eccc-date-lag}

```{r heat-eccc-date-lag}
expand_months <- function(d, hospital_id, survey_month, survey_year, station_id) {
  data.table(
    hospital_id = hospital_id,
    survey_year = survey_year,
    survey_month = survey_month,
    station_id = station_id,
    offset = 0:24,
    date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
  )
}

#apply the function
#identifies how many months a given month is offset from the 0th month for that surveys survey date
survey_sf$dates <- ymd(paste(survey_sf$survey_year, survey_sf$survey_month, "01", sep = "-")) + months(1) - days(1)
#FOR EVERY UNIQUE survey AND SURVEY DATE COMBINATION <- this is key
surveys_un <- unique(survey_sf, by = c("survey_id","dates"))
surveys_un <- as.data.table(surveys_un)
survey_expanded <- surveys_un[, expand_months(dates, hospital_id, survey_month, survey_year, station_id), by = .(hospital_id, dates)]
survey_expanded <- survey_expanded %>% select(-3)
survey_expanded_heat_eccc <- survey_expanded %>% rename(year_month = date)
```

#### Date Join {#heat-eccc-date-join}

Create heat event data for all months before narrowing the range and then join with the expanded days.

```{r heat-eccc-date-join}
#arrange the data chronologically once again and setup columns for joining, add another heat classifier since this is being grouped to months (month has_heat)
heat_monthly_eccc <- heat_flagged_eccc %>%
  st_drop_geometry() %>% #drop the geometry for faster speeds
  arrange(station_id, date) %>%
  mutate(year_month = as.integer(format(date, "%Y%m"))) %>%
  group_by(hospital_id, year_month) %>%
  summarise(
    has_heat = any(heat_event, na.rm = TRUE),                     
    .groups = "drop"
  ) %>%
  mutate(
    year  = as.integer(year_month / 100), #need a year eventually
    month = as.integer(year_month %% 100), #need a month eventually
    year_month = as.character(year_month)
  )
  

#join with the survey data
survey_with_heat_eccc <- survey_expanded_heat_eccc %>% #select(-5,-6) %>% 
  left_join(heat_monthly_eccc, 
            by = c("hospital_id", "year_month")) %>%
  #rename(year_month=dates) %>% 
  mutate(
    heat_flag = as.integer(coalesce(has_heat, FALSE))) #integer based has_heat           

heat_month_zero_daily_eccc <- filter(survey_with_heat_eccc, survey_year==year & survey_month==month) %>% st_drop_geometry()

```

#### Matrix {#heat-matrix-eccc}

`survey_with_heat_eccc` has all the interesting information in it.

```{r heat-matrix-eccc}
heat_summary_matrix_eccc <- survey_with_heat_eccc %>% 
  filter(heat_flag == 1) %>% #only the months with floods in them
  group_by(hospital_id, survey_year, survey_month, offset) %>% #group by the offsets
  slice_head(n=1) %>%
  right_join(survey_expanded_heat_eccc, by=c("hospital_id", "survey_year", "survey_month", "offset")) %>%
  mutate(heat_flag = replace_na(heat_flag, 0)) %>% #make sure there are no NAs
  ungroup() %>%
  as.data.table() %>%
  left_join(survey_sf %>% select(hospital_id, station_id, nearest_dist), by='hospital_id') %>%
  dcast(.,
      hospital_id + survey_month + survey_year + station_id + nearest_dist~ paste0("month_", offset),
      value.var = "heat_flag",
      fun.aggregate = max
    ) %>%
  setcolorder(c("hospital_id", "survey_month", "survey_year", 'station_id', 'nearest_dist', paste0("month_", 0:24))) 

write.csv(heat_summary_matrix_eccc, r"(./data/heat_matrix_eccc.csv)")
```

### Heat Days {#heat-daily-arb}

```{r heat-combined-daily-arb}
combined_hourly_daily <- read_parquet(r"(./data/odhf_weather_stations.parquet)")
temp_threshold <- 30
consecutive_days <- 3

heat_flagged <- combined_hourly_daily %>%
  arrange(station_id, date) %>% #arrange by station and chronologically
  unique(.,by=c("station_id", "date")) %>% #filter by only unqiue stations and dates, can't have any duplicates
  group_by(station_id) %>% #applying only to stations
  mutate(
    hot = max_temp >= temp_threshold, #it is hot if it is greater than the arbitrary threshold               
    run_id = rleid(hot) #group chronological runs together
  ) %>%
  group_by(station_id, run_id) %>% #group the runs together
  mutate(
    run_len = sum(hot), #sum the number of hot events per run 
    heat_event = hot & run_len >= consecutive_days    
  ) %>%
  ungroup() %>%
  select(-run_id, -run_len) %>% 
  mutate(date = as.Date(date))

heat_flagged_arb <- heat_flagged
```

#### Date Lag {#heat-arb-date-lag}

Run the same date lags as with other hazards.

```{r heat-arb-date-lag}
expand_months <- function(d, hospital_id, survey_month, survey_year, station_id) {
  data.table(
    hospital_id = hospital_id,
    survey_year = survey_year,
    survey_month = survey_month,
    station_id = station_id,
    offset = 0:24,
    date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
  )
}

#apply the function
#identifies how many months a given month is offset from the 0th month for that surveys survey date
survey_sf$dates <- ymd(paste(survey_sf$survey_year, survey_sf$survey_month, "01", sep = "-")) + months(1) - days(1)
#FOR EVERY UNIQUE survey AND SURVEY DATE COMBINATION <- this is key
surveys_un <- unique(survey_sf, by = c("survey_id","dates"))
surveys_un <- as.data.table(surveys_un)
survey_expanded <- surveys_un[, expand_months(dates, hospital_id, survey_month, survey_year, station_id), by = .(hospital_id, dates)]
survey_expanded <- survey_expanded %>% select(-3)
survey_expanded_heat_days <- survey_expanded %>% rename(year_month = date)
```

#### Date Join {#heat-arb-date-join}

Create heat event data for all months before narrowing the range and then join with the expanded days.

```{r heat-arb-date-join}
heat_monthly <- heat_flagged %>%
  arrange(station_id, date) %>% #arrange chronologically
  mutate(year_month = as.integer(format(date, "%Y%m"))) %>%        
  group_by(station_id, year_month) %>% #group by year_month and station
  summarise(
    has_heat = any(heat_event, na.rm = TRUE), #we only care about any heat event in the month
    .groups = "drop"
  ) %>%
  mutate(
    year  = as.integer(year_month / 100), #need a year eventually
    month = as.integer(year_month %% 100), #need a month eventually
    year_month = as.character(year_month)
  )

survey_with_heat_days <- survey_expanded_heat_days %>% #join with monthly data
  left_join(heat_monthly, 
            by = c("station_id", "year_month")) %>%          
  mutate(
    heat_flag = as.integer(coalesce(has_heat, FALSE))) #coalesce to an integer
  
heat_flagged_daily_normalized <- heat_flagged %>% #normalized is a misnomer, we are just looking for the month 0 stuff with this one
  mutate(year_month = as.integer(format(date, "%Y%m"))) %>% 
  mutate(
    year  = as.integer(year_month / 100),      
    month = as.integer(year_month %% 100),
    year_month = as.character(year_month)
  ) %>% left_join(survey_expanded_heat_days, by = c("station_id","year_month")) %>% 
  filter(offset==0, heat_event == TRUE)
```

#### Matrix {heat-arb-matrix}

Create the summary matrix of 1's and 0's.

```{r heat-arb-matrix}
heat_summary_matrix_arb <- survey_with_heat_days %>% 
  filter(heat_flag == 1) %>% #only the months with floods in them
  group_by(hospital_id, survey_year, survey_month, offset) %>% #group by the offsets
  slice_head(n=1) %>%
  right_join(survey_expanded_heat_days, by=c("hospital_id", "survey_year", "survey_month", "offset")) %>%
  mutate(heat_flag = replace_na(heat_flag, 0)) %>% #make sure there are no NAs
  ungroup() %>%
  as.data.table() %>%
  left_join(survey_sf %>% select(hospital_id, station_id, nearest_dist), by='hospital_id') %>%
  dcast(.,
      hospital_id + survey_month + survey_year + station_id + nearest_dist~ paste0("month_", offset),
      value.var = "heat_flag",
      fun.aggregate = max
    ) %>%
  setcolorder(c("hospital_id", "survey_month", "survey_year", 'station_id', 'nearest_dist', paste0("month_", 0:24))) 

write.csv(heat_summary_matrix_arb, r"(./data/heat_matrix_dayCriteria.csv)")
```


## PM 2.5 {#pm25}

### Environmental Data Preparation {#pm25-data-preparation}

see the index.

```{r pm25-data-cleaning, warning=FALSE}

odhf_pm25 <- odhf_valid %>%
  mutate(
    pm25_start = make_date(survey_year - 2, survey_month, 1),  #start 2 years before
    pm25_end = make_date(survey_year, survey_month, days_in_month(make_date(survey_year, survey_month, 1))))  #last day of the month

head(odhf_pm25)

#define years
years <- 2015:2018

#define the base path
base_path <- "./data/pm25/"

#read, standardize, and pivot dataset
load_pm25_data <- function(year) {
  suffix <- substr(as.character(year), 3, 4) #get just the year from the year range
  file   <- file.path(base_path, paste0("aqsmk_avc_", suffix, ".csv")) #the base path of the files + the year
  df <- read_csv(file, show_col_types = FALSE)
  names(df)[1] <- "postalcode"                        
  df_long <- df %>% #pivot the data to long format based on the columns in the data
    pivot_longer(
      cols      = starts_with("aqsmk"), #column for the aqsmk data 
      names_to  = "var",
      values_to = "pm25"
    ) %>%
    mutate(
      ym_raw = str_extract(var, "\\d{2}_\\d{2}"), #regex
      year   = 2000 + as.integer(substr(ym_raw, 1, 2)),  
      month  = as.integer(substr(ym_raw, 4, 5)) #all data points are january     
    ) %>%
    select(postalcode, province, year, month, pm25) %>%
    filter(!is.na(pm25)) %>% rename(postcode=postalcode)
  
  df_long
}

pm25_all_years <- map_dfr(years, load_pm25_data) #load all the csvs
#
##turn month into a factor with correct order
#pm25_all_years <- pm25_all_years %>%
#  mutate(month = factor(month, levels = c("jan", "feb", "mar", "apr", "may", "jun",
#                                          "jul", "aug", "sep", "oct", "nov", "dec")))

pm25_all_years <- pm25_all_years %>% rename(survey_year=year) %>% rename(survey_month=month)

head(pm25_all_years)
```

### Date Lag {pm25-date-lag}

```{r pm-date-lag}
expand_months <- function(d, hospital_id, survey_month, survey_year, postalcode) {
  data.table(
    hospital_id = hospital_id,
    survey_year = survey_year,
    survey_month = survey_month,
    postcode = postalcode,
    offset = 0:24,
    date = format(ceiling_date(d %m-% months(0:24), "month") - days(1), "%Y%m")
  )
}

#apply the function
#identifies how many months a given month is offset from the 0th month for that surveys survey date
odhf_pm25$dates <- ymd(paste(odhf_pm25$survey_year, odhf_pm25$survey_month, "01", sep = "-")) + months(1) - days(1)
#FOR EVERY UNIQUE survey AND SURVEY DATE COMBINATION <- this is key
surveys_un <- unique(odhf_pm25, by = c("survey_id","dates"))
surveys_un <- as.data.table(surveys_un)
survey_expanded <- surveys_un[, expand_months(dates, hospital_id, survey_month, survey_year, postalcode), by = .(hospital_id, dates)]
survey_expanded <- survey_expanded %>% select(-3)
survey_expanded_pm25 <- survey_expanded %>% rename(year_month = date)


```


### Data Manipulation {pm25-data-manipulation}

```{r pm25-data-manipulation}
#turn month into a factor with correct order
pm25_all_years <- pm25_all_years %>%
  select(postcode, province, survey_year, survey_month, pm25)



#join with the expanded survey info
pm25_survey <- survey_expanded_pm25 %>% 
  left_join(
    pm25_all_years %>% select(postcode, survey_year, survey_month, pm25),
    by = c("postcode", "survey_year", "survey_month")
  ) %>%
  group_by(hospital_id, survey_year) %>%
  mutate(pm25 = if (all(is.na(pm25))) NA_real_ else first(na.omit(pm25))) %>%
  ungroup()

#summarise into the matrix of 1's and 0's
pm25_summary_matrix <- as.data.table(pm25_survey) %>%
  dcast(
    hospital_id + survey_month + survey_year ~ paste0("month_", offset),
    value.var = "pm25",
    fun.aggregate = first
  ) %>%
  setcolorder(c("hospital_id", "survey_month", "survey_year", paste0("month_", 0:24)))

#tons of NA's in this dataset
write.csv(pm25_summary_matrix, "./data/pm25_presence_absence.csv")


```

# Appendix

## Survey Data Generation

```{r sample-data-generation, eval=FALSE}
#TODO incorporate the reference for this data into the index
#generation of data based on entries in the hospital data 
#if there are 20 hospitals, make 20 random months for survey between 1 and 12 and random years choosing between 2016 and 2020
#https://www150.statcan.gc.ca/n1/en/pub/13-26-0001/2020001/ODHF_v1.1.zip
#make sure the latitude and longitude columns exist and that there is an id column as well
#set a random seed
set.seed(933)
#read in the raw data
odhf <- read.csv("./data/odhf_v1.1.csv")
#create an id list
odhf$hospital_id <- 1:nrow(odhf)
#can only use valid lat/long locations
odhf_valid <- odhf[!is.na(odhf[["latitude"]]) & 
                             !is.na(odhf[["longitude"]]), ]
odhf_valid <- odhf_valid %>% filter(odhf_facility_type=="Hospitals") %>%  select(c("latitude", "longitude", "facility_name", "odhf_facility_type", "hospital_id", "postal_code"))
odhf_valid$postal_code <- gsub(" ", "", odhf_valid$postal_code)
odhf_valid <- odhf_valid %>% rename(postalcode = postal_code)
odhf_valid[[survey_month_col]] <- sample(1:12, size = nrow(odhf_valid), replace = TRUE)
odhf_valid[[survey_year_col]] <- sample(2018:2021, size=nrow(odhf_valid), replace = TRUE)
setnames(odhf_valid, c("latitude", "longitude"), c("lat","long"))
write.csv(odhf_valid, "./data/odhf_hospitals.csv")
```
